---
title: "cfDNA fragment size distribution"
author: "Darren Liu"
date: "2024-04-01"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = "W:/Shared/LRI/Labs/engclab/LAB_MEMBER_CENTRAL/Darren_Liu/3_cf-DNA")
```

# Load library
```{r libraries}
library(tidyverse)
library(ggpubr)
library(ggplot2)
library(scales)
library(viridis)
```

# Import Data
```{r import}

# Import frag/bin and meta data  
df.frags <- readRDS("protocol_ms/data/11-combine_bin/binfrag_summary.rds")
meta <- read.csv("protocol_ms/data/phtscfx_meta240304.csv")
```

# Clean and reshape data
```{r clean}

# Summarise counts for each cfDNA fragment size per sample
df.frags <- df.frags %>%
  group_by(id) %>% 
  summarise(
    across(counts_100:counts.corrected_638, sum)
  )

# Merge with meta data 
df.frags <- inner_join(meta, df.frags, by="id") 

# Uncorrected counts
df.frags1 <- df.frags %>% 
  select(matches("^id$|^cancer$|^group2$|^group3$|^ca_1st$|^ca_2nd$|^counts_[0-9]+$"))

# Reshape  
df.frags1 <- df.frags1 %>%
  pivot_longer(cols = starts_with("counts_"), 
               names_to = "frag_size",
               values_to = "counts") %>% 
  mutate(frag_size = as.numeric(str_remove(frag_size, "counts_")))

# Corrected counts 
df.frags1.corrected <- df.frags %>% 
  select(matches("^id$|^cancer$|^group2$|^group3$|^ca_1st$|^ca_2nd$|^counts\\.corrected_[0-9]+$"))

# Reshape dataframe 
df.frags1.corrected <- df.frags1.corrected %>%
  pivot_longer(cols = starts_with("counts.corrected_"), 
               names_to = "frag_size",
               values_to = "counts.corrected") %>% 
  mutate(frag_size = as.numeric(str_remove(frag_size, "counts.corrected_")))

# Merge uncorrected/corrected counts 
df.frags2 <- df.frags1 %>% 
  full_join(df.frags1.corrected, 
            by = c("id", "cancer", "group2", "group3", "ca_1st", "ca_2nd", "frag_size"))


# calculate frequency and total counts
df.frags2 <- df.frags2 %>%
  group_by(id) %>% 
  mutate(
    total.counts = sum(counts), 
    total.counts.corrected = sum(counts.corrected),
    freq = counts / total.counts,
    freq.corrected = counts.corrected/total.counts.corrected
  ) %>% ungroup() %>%
  mutate(
    frac = case_when(
        frag_size >= 100 & frag_size <=250 ~ "Mononucleosome", 
        frag_size >= 251 & frag_size <=450 ~ "Dinucleosome", 
        frag_size >= 451 & frag_size <=650 ~ "Trinucleosome"
  )) %>%
  mutate(
    frac = factor(frac, levels = c("Mononucleosome", "Dinucleosome", "Trinucleosome"))
  ) %>% 
  relocate(frac, .before = frag_size)

# Save files
write.csv(df.frags2, "protocol_ms/data/12-size_distr/12-size_distr_summary240403.csv", row.names = FALSE)
saveRDS(df.frags2, "protocol_ms/data/12-size_distr/size_distr.rds")
```

# Figure 1A-C. Fragment Size Distribution by Cancer Status 
```{r fig1_ca}

# Calculate median by cancer status 
df.frags.ca <- df.frags2 %>% ungroup() %>% 
  group_by(cancer, frag_size, frac) %>%
  summarise(
    median.freq = median(freq),
    median.freq.correct = median(freq.corrected)
  ) %>%
  mutate(
    cancer = case_when(
      cancer == "Yes" ~ "Cancer",
      cancer == "No" ~ "No Cancer")
  )

# Sample sizes (used for labeling)
n_cancer <- 49
n_no_cancer <- 50

# Update group names with sample sizes
df.frags.ca1 <- df.frags.ca %>%
  mutate(
    cancer = recode(cancer,
                    `Cancer` = paste0("Cancer (n = ", n_cancer, ")"),
                    `No Cancer` = paste0("No Cancer (n = ", n_no_cancer, ")")),)
  
# Select color palette
viridis_palette <- viridis(30)
custom_color <- viridis_palette[c(1, 29)]
show_col(viridis_palette)

# Plot distribution for each nucleosome fraction
p1.ca <- df.frags.ca1 %>%
  group_by(frac) %>% 
  group_map(~ggplot(.) + 
            aes(x = frag_size, y = median.freq.correct, color = cancer) +
            scale_color_manual(name = "Group",
                               values = custom_color) +
            geom_line(size = 0.75) + 
            labs(x = "Fragment size (bp)", y = "Frequency", title = .y[[1]]) + 
            theme(axis.title.x = element_text(size = 10),
                  axis.title.y = element_text(size = 10),
                  plot.title = element_text(size = 10)))

# Combine plots 
p1.ca.combined <- ggarrange(plotlist = p1.ca, 
                             ncol = 3, nrow = 1, labels = c("A", "B", "C"), common.legend = TRUE, legend = "right")

# Save as tiff
imagedir="protocol_ms/data/12-size_distr"
out.plot <- file.path(imagedir, "fragdist_ca_240404.tiff")
ggsave(filename = out.plot, plot = p1.ca.combined, device = "tiff", dpi = 300, width = 8, height = 3)
```

# Figure 1D-F. Fragment Size Distribution by SMN Status
```{r fig1_smn}

# Calculate median by SMN status
df.frags.smn <- df.frags2 %>% ungroup() %>% 
  group_by(group2, frag_size, frac) %>%
  summarise(
    median.freq = median(freq),
    median.freq.correct = median(freq.corrected)
  ) %>% 
  mutate(
    group2 = factor(group2, levels = c("SMN", "PMN", "No Cancer")),
  )

# Sample sizes (used for labeling)
n_smn <- 26
n_pmn <- 23
n_no_cancer <- 50

# Update group names with sample sizes
df.frags.smn1 <- df.frags.smn %>%
  mutate(
    group2 = recode(group2,
                    `SMN` = paste0("SMN (n = ", n_smn, ")"),
                    `PMN` = paste0("PMN (n = ", n_pmn, ")"),
                    `No Cancer` = paste0("No Cancer (n = ", n_no_cancer, ")")),)
  
# Select color palette
viridis_palette <- viridis(30)
custom_color <- viridis_palette[c(1, 16, 29)]
show_col(viridis_palette)

# Plot distribution for each nucleosome fraction
p1.smn <- df.frags.smn1 %>%
  group_by(frac) %>% 
  group_map(~ggplot(.) + 
            aes(x = frag_size, y = median.freq.correct, color = group2) +
            geom_line(size = 0.75) + 
            labs(x = "Fragment size (bp)", y = "Frequency", title = .y[[1]]) +
            scale_color_manual(name = "Group",
                               values = custom_color) +
            theme(axis.title.x = element_text(size = 10),
                  axis.title.y = element_text(size = 10),
                  plot.title = element_text(size = 10)))


# Combine plots
p1.smn.combined <- ggarrange(plotlist = p1.smn, 
                             ncol = 3, nrow = 1, labels = c("D", "E", "F"), common.legend = TRUE, legend = "right")

# Save as tiff
imagedir="protocol_ms/data/12-size_distr"
out.plot <- file.path(imagedir, "fragdist_smn_240404.tiff")
ggsave(filename = out.plot, plot = p1.smn.combined, device = "tiff", dpi = 300, width = 8, height = 3)
```

# Figure 1G-I. Fragment Size Distribution by SMN Status and Plasma Draw
```{r fig1_pre}

# Calculate median by SMN and plasma draw time
df.frags.pre <- df.frags2 %>% ungroup() %>% 
  group_by(group3, frag_size, frac) %>%
  summarise(
    median.freq = median(freq),
    median.freq.correct = median(freq.corrected)
  ) %>% 
  mutate(
    group3 = factor(group3, levels = c("Pre-SMN", "Post-SMN", "Pre-PMN", "Post-PMN", "No Cancer"))
  )

# Sample sizes (used for labeling)
n_presmn <- 4
n_postsmn <- 22
n_prepmn <- 5
n_postpmn <- 18
n_no_cancer <- 50

# Update group names with sample sizes
df.frags.pre1 <- df.frags.pre %>%
  mutate(group3 = recode(group3,
                         `Pre-SMN` = paste0("Pre-SMN (n = ", n_presmn, ")"),
                         `Pre-PMN` = paste0("Pre-PMN (n = ", n_prepmn, ")"),
                         `Post-SMN` = paste0("Post-SMN (n = ", n_postsmn, ")"),
                         `Post-PMN` = paste0("Post-PMN (n = ", n_postpmn, ")"),
                         `No Cancer` = paste0("No Cancer (n = ", n_no_cancer, ")")))

# Select color palette
viridis_palette <- viridis(30)
custom_color <- viridis_palette[c(1, 9, 16, 24, 29)]
show_col(viridis_palette)

# Plot distribution for each nucleosome fraction
p1.pre <- df.frags.pre1 %>%
  group_by(frac) %>% 
  group_map(~ggplot(.) + 
            aes(x = frag_size, y = median.freq.correct, color = group3) +
            geom_line(size = 0.75) + 
            labs(x = "Fragment size (bp)", y = "Frequency", title = .y[[1]]) +
            scale_color_manual(name = "Group",
                               values = custom_color) +
            theme(axis.title.x = element_text(size = 10),
                  axis.title.y = element_text(size = 10),
                  plot.title = element_text(size = 10)))


# Combine plots
p1.pre.combined <- ggarrange(plotlist = p1.pre, 
                             ncol = 3, nrow = 1, labels = c("D", "E", "F"), common.legend = TRUE, legend = "right")

# Save as tiff
imagedir="protocol_ms/data/12-size_distr"
out.plot <- file.path(imagedir, "fragdist_pre_240404.tiff")
ggsave(filename = out.plot, plot = p1.pre.combined, device = "tiff", dpi = 300, width = 8, height = 3)

```

# Combine and save main figures
```{r main_fig_save}

# Combine all size distribution plots
p1.combined <- ggarrange(p1.ca.combined, p1.smn.combined, p1.pre.combined, ncol = 1, nrow = 3, common.legend = F, legend = "right")
imagedir="protocol_ms/data/12-size_distr"
out.plot <- file.path(imagedir, "Figure1_fragdist_combined_240404.tiff")
ggsave(filename = out.plot, plot = p1.combined, device = "tiff", dpi = 300, width = 8, height = 9)

```

# Statistical Analysis, KS Test - All Nucleosomes
```{r ks_res_overall}

# Function to perform pairwise KS tests
ks_test <- function(data, variable_name) {
  # Get the list of groups
  groups <- unique(data$group)

  # Generate all pairs of groups
  pairs <- combn(groups, 2, simplify = FALSE)

  # Perform the KS test for each pair
  results <- lapply(pairs, function(pair) {
    group1_data <- data$value[data$group == pair[1]]
    group2_data <- data$value[data$group == pair[2]]
    ks_result <- ks.test(group1_data, group2_data)

    # Return a data frame with the test result and the groups compared
    data.frame(
      variable = variable_name,
      group1 = pair[1],
      group2 = pair[2],
      statistic = ks_result$statistic,
      p.value = ks_result$p.value
    )
  })

  # Combine the results into a single data frame
  do.call(rbind, results)
}

# Reshape dataframes 
df.frags.ca.long <- df.frags.ca %>%
  select(cancer, median.freq.correct, median.freq, frac, frag_size) %>%
  pivot_longer(cols = c("median.freq.correct", "median.freq"), names_to = "variable", values_to = "value") %>%
  rename(group = cancer)
 
df.frags.smn.long <- df.frags.smn %>%
  select(group2, median.freq.correct, median.freq, frac, frag_size) %>%
  pivot_longer(cols = c("median.freq.correct", "median.freq"), names_to = "variable", values_to = "value") %>%
  rename(group = group2)
 
df.frags.pre.long <- df.frags.pre %>%
  select(group3, median.freq.correct, median.freq, frac, frag_size) %>%
  pivot_longer(cols = c("median.freq.correct", "median.freq"), names_to = "variable", values_to = "value") %>%
  rename(group = group3)

# Group the data by the variable and split it into a list of data frames
df.grouped.ca <- split(df.frags.ca.long, df.frags.ca.long$variable)
df.grouped.smn <- split(df.frags.smn.long, df.frags.smn.long$variable)
df.grouped.pre <- split(df.frags.pre.long, df.frags.pre.long$variable)

# Apply the ks test to each group
ks.res.ca <- lapply(names(df.grouped.ca), function(name) ks_test(df.grouped.ca[[name]], name))
ks.res.smn <- lapply(names(df.grouped.smn), function(name) ks_test(df.grouped.smn[[name]], name))
ks.res.pre <- lapply(names(df.grouped.pre), function(name) ks_test(df.grouped.pre[[name]], name))

# Combine list (non-Gc and Gc-corrected results) into single dataframe 
df.res.ca <- do.call(rbind, ks.res.ca)
df.res.smn <- do.call(rbind, ks.res.smn)
df.res.pre <- do.call(rbind, ks.res.pre)
df.res.all <- rbind(df.res.ca, df.res.smn, df.res.pre)

# Save CSV
write.csv(df.res.all, "protocol_ms/data/12-size_distr/kstest_overall_pairwise240404.csv", row.names = FALSE)

```

# Statistical Analysis, KS Test - Individual Nucleosomes
```{r ks_res_indv}

## Functions
# Split data
split_data <- function(df) {
  split(df, list(df$variable, df$frac))
}

# Clean up results
process_res <- function(df) {
  df %>%
    mutate(frac = str_extract(variable, "\\.[^.]+$")) %>%
    mutate(frac = str_remove(frac, "\\."), 
           variable = str_remove(variable, "\\.[^.]+$")) %>%
    relocate(variable, .after = group2) %>%
    relocate(frac, .after = variable)
}

# Create list from results for each subgroup 
df.frags.all.long <- list(ca = df.frags.ca.long, smn = df.frags.smn.long, pre = df.frags.pre.long)

results <- list()
# Loop through list and apply functions
for (name in names(df.frags.all.long)) {
  group <- split_data(df.frags.all.long[[name]])
  ks.res <- lapply(names(group), function(group_name) ks_test(group[[group_name]], group_name))
  res <- do.call(rbind, ks.res)
  df.res <- process_res(res)
  results[[name]] <- df.res
}

# Combine list into single dataframe
df.res.all <- rbind(results$ca, results$smn, results$pre)

# Save CSV
write.csv(df.res.all, "protocol_ms/data/12-size_distr/kstest_frac_pairwise240404.csv", row.names = FALSE)

```

# SFigure 1. Cumulative Frequency Distribution 
```{r cum_freq}

# Calculate cumulative frequency 
calculate_cum_freq <- function(df, group_var) {
  df %>%
    group_by(group = {{group_var}}, frac, variable) %>%
    arrange(frag_size) %>%
    mutate(cum_freq = cumsum(value)) %>%
    ungroup()
}

# Plot cumulative frequency 
plot_cumulative_frequency <- function(df, color_palette) {
  p <- df %>%
    filter(variable != "median.freq") %>%
    group_by(frac, variable) %>%
    group_map(~ggplot(data = .x, aes(x = frag_size, y = cum_freq, color = group)) +
                geom_line(size = 0.75) +
                labs(x = "Fragment size (bp)", y = "Cumulative Frequency", title = .y[[1]]) +
                scale_color_manual(name = "Group", values = color_palette) +
                theme(axis.title.x = element_text(size = 10),
                      axis.title.y = element_text(size = 10),
                      plot.title = element_text(size = 10)))
  return(p)
}


# Calculate cumlative frequency for subgroups
df.cumfreq.ca <- calculate_cum_freq(df.frags.ca.long, group)
df.cumfreq.smn <- calculate_cum_freq(df.frags.smn.long, group)
df.cumfreq.pre <- calculate_cum_freq(df.frags.pre.long, group)

# Rename for figure labeling
df.cumfreq.smn <- df.cumfreq.smn %>%
  mutate(
    group = recode(group,
                    `SMN` = paste0("SMN (n = ", n_smn, ")"),
                    `PMN` = paste0("PMN (n = ", n_pmn, ")"),
                    `No Cancer` = paste0("No Cancer (n = ", n_no_cancer, ")")),)


df.cumfreq.pre <- df.cumfreq.pre %>%
  mutate(group = recode(group,
                         `Pre-SMN` = paste0("Pre-SMN (n = ", n_presmn, ")"),
                         `Pre-PMN` = paste0("Pre-PMN (n = ", n_prepmn, ")"),
                         `Post-SMN` = paste0("Post-SMN (n = ", n_postsmn, ")"),
                         `Post-PMN` = paste0("Post-PMN (n = ", n_postpmn, ")"),
                         `No Cancer` = paste0("No Cancer (n = ", n_no_cancer, ")")))


# Define color palette
viridis_palette <- viridis(30)
show_col(viridis_palette)
color_ca <- viridis_palette[c(1, 29)]
color_smn <- viridis_palette[c(1, 16, 29)]
color_pre <- viridis_palette[c(1, 9, 16, 24, 29)]

# Generate plot for subgroups
p1.cf.ca <- plot_cumulative_frequency(df.cumfreq.ca, color_ca)
p1.cf.smn <- plot_cumulative_frequency(df.cumfreq.smn, color_smn)
p1.cf.pre <- plot_cumulative_frequency(df.cumfreq.pre, color_pre)
  

# Combine plots
p1.cf.ca.combined <- ggarrange(plotlist = p1.cf.ca, ncol = 3, nrow = 1, labels = c("A", "B", "C"), common.legend = TRUE, legend = "right")

p1.cf.smn.combined <- ggarrange(plotlist = p1.cf.smn, ncol = 3, nrow = 1, labels = c("D", "E", "F"), common.legend = TRUE, legend = "right")

p1.cf.pre.combined <- ggarrange(plotlist = p1.cf.pre, ncol = 3, nrow = 1, labels = c("G", "H", "I"), common.legend = TRUE, legend = "right")


# Combine all and save  
p1.cf.combined <- ggarrange(p1.cf.ca.combined, p1.cf.smn.combined, p1.cf.pre.combined, ncol = 1, nrow = 3, common.legend = F, legend = "right")

imagedir="protocol_ms/data/12-size_distr"
out.plot <- file.path(imagedir, "SFigure1_cumfreq_240416.tiff")
ggsave(filename = out.plot, plot = p1.cf.combined, device = "tiff", dpi = 300, width = 8, height = 9, bg = "white")


```

# SFigure 2. Difference in Fragment Size Frequency 
```{r delta_f}

# Function to calculate delta F
calculate_delta_f <- function(df, pairs) {
  # Reshape to wide format to facilitate group-wise subtraction
  df_wide <- df %>%
    pivot_wider(names_from = group, values_from = value)

  # Calculate delta_f for each pair specified in pairs
  for (pair in pairs) {
    group1 <- gsub("[ -]", "_", pair[1])  # Replace spaces and hyphens with underscores
    group2 <- gsub("[ -]", "_", pair[2])  
    delta_col_name <- paste0("delta_f_", group1, "_", group2)
    
    # Calculate difference
    df_wide <- df_wide %>%
      mutate(!!delta_col_name := !!sym(pair[1]) - !!sym(pair[2]))
  }

  # Reshape to long format
  df_long <- df_wide %>%
    pivot_longer(
      cols = starts_with("delta_f"),
      names_to = "group",
      values_to = "delta_f"
    )

  return(df_long)
}

# Define pairs 
pairs_ca <- list(c("Cancer", "No Cancer"))
pairs_smn <- list(c("SMN", "No Cancer"), c("PMN", "No Cancer"))
pairs_pre <- list(c("Pre-SMN", "No Cancer"), c("Post-SMN", "No Cancer"), c("Pre-PMN", "No Cancer"), c("Post-PMN", "No Cancer"))

# Calculate delta F for subgroups 
df.delta_f.ca <- calculate_delta_f(df.frags.ca.long, pairs_ca)
df.delta_f.smn <- calculate_delta_f(df.frags.smn.long, pairs_smn)
df.delta_f.pre <- calculate_delta_f(df.frags.pre.long, pairs_pre)

# Define color palette
viridis_palette <- viridis(30)
show_col(viridis_palette)
color_ca <- viridis_palette[1]
color_smn <- viridis_palette[c(1, 16)]
color_pre <- viridis_palette[c(1, 9, 16, 24)] 

# Rename group column for figure labeling
df.delta_f.ca$group <- recode(df.delta_f.ca$group,
                                delta_f_Cancer_No_Cancer = "Cancer (n = 49)")

df.delta_f.smn$group <- recode(df.delta_f.smn$group,
                                delta_f_SMN_No_Cancer = "SMN (n = 26)",
                                delta_f_PMN_No_Cancer = "PMN (n = 23)") 

df.delta_f.pre$group <- recode(df.delta_f.pre$group,
                                delta_f_Pre_SMN_No_Cancer = "Pre-SMN (n = 4)", 
                                delta_f_Pre_PMN_No_Cancer = "Pre-PMN (n = 5)",
                                delta_f_Post_SMN_No_Cancer = "Post-SMN (n = 22)",
                                delta_f_Post_PMN_No_Cancer = "Post-PMN (n = 18)") 

# Reorder factors for labeling
df.delta_f.smn <- df.delta_f.smn %>% mutate(group = factor(group, levels = c("SMN (n = 26)", "PMN (n = 23)")))
df.delta_f.pre <- df.delta_f.pre %>% mutate(group = factor(group, levels = c("Pre-SMN (n = 4)", "Post-SMN (n = 22)", "Pre-PMN (n = 5)", "Post-PMN (n = 18)")))

# Function to plot delta F  
plot_delta_f <- function(df, color_palette) {
  p <- df %>%
    filter(variable != "median.freq") %>%
    group_by(frac, variable) %>%
    group_map(~ggplot(data = .x, aes(x = frag_size, y = delta_f, color = group)) +
                geom_line() +
                labs(x = "Fragment size (bp)", y = expression(paste(Delta, "F")), title = .y[[1]]) +
                scale_color_manual(name = "Group", values = color_palette) +
                theme(axis.title.x = element_text(size = 10),
                      axis.title.y = element_text(size = 10),
                      plot.title = element_text(size = 10)))
  return(p)
}

# Generate plot for subgroups
p1.delta_f.ca <- plot_delta_f(df.delta_f.ca, color_ca)
p1.delta_f.smn <- plot_delta_f(df.delta_f.smn, color_smn)
p1.delta_f.pre <- plot_delta_f(df.delta_f.pre, color_pre)

# Combine plots
p1.delta_f.ca.combined <- ggarrange(plotlist = p1.delta_f.ca, ncol = 3, nrow = 1, labels = c("A", "B", "C"), common.legend = TRUE, legend = "bottom")

p1.delta_f.smn.combined <- ggarrange(plotlist = p1.delta_f.smn, ncol = 3, nrow = 1, labels = c("D", "E", "F"), common.legend = TRUE, legend = "bottom")

p1.delta_f.pre.combined <- ggarrange(plotlist = p1.delta_f.pre, ncol = 3, nrow = 1, labels = c("G", "H", "I"), common.legend = TRUE, legend = "bottom")

# Combine all and save  
p1.delta_f.combined <- ggarrange(p1.delta_f.ca.combined, p1.delta_f.smn.combined, p1.delta_f.pre.combined, ncol = 1, nrow = 3, common.legend = F, legend = "bottom")

imagedir="protocol_ms/data/12-size_distr"
out.plot <- file.path(imagedir, "SFigure2_delta_f_240416.tiff")
ggsave(filename = out.plot, plot = p1.delta_f.combined, device = "tiff", dpi = 300, width = 8, height = 9, bg = "white")

```

# SFigure 3. Difference in Cumulative Frequency Distribution
```{r cum_freq_delta}

# FUnction to calculate total freq frequency 
calculate_total_freq <- function(df, group_var) {
  df %>%
    group_by(group = {{group_var}}, variable) %>%
    summarize(total_freq = sum(value), .groups = 'drop')
}

# Calculating total frequency 
df.total.freq.ca <- calculate_total_freq(df.frags.ca.long, group)
df.total.freq.smn <- calculate_total_freq(df.frags.smn.long, group)
df.total.freq.pre <- calculate_total_freq(df.frags.pre.long, group)

# Function to normalize cumulative frequency 
calculate_norm_cumfreq <- function(df, total_freq_df) {
  df_cumfreq_norm <- df %>%
    left_join(total_freq_df, by = c("group", "variable")) %>%
    # Calculate the normalized frequency for each fragment size within each group
    mutate(norm_value = value / total_freq) %>%
    arrange(frag_size) %>%
    group_by(group, frac, variable) %>%
    # Calculate the normalized cumulative frequency
    mutate(norm_cum_freq = cumsum(norm_value)) %>%
    select(-total_freq, -norm_value) %>% # Unncessary columns
    ungroup()

  return(df_cumfreq_norm)
}

# Calculate normalized cumulative frequency 
df.cumfreq_norm.ca <- calculate_norm_cumfreq(df.frags.ca.long, df.total.freq.ca)
df.cumfreq_norm.smn <- calculate_norm_cumfreq(df.frags.smn.long, df.total.freq.smn)
df.cumfreq_norm.pre <- calculate_norm_cumfreq(df.frags.pre.long, df.total.freq.pre)

# Define pairs 
pairs_ca <- list(c("Cancer", "No Cancer"))
pairs_smn <- list(c("SMN", "No Cancer"), c("PMN", "No Cancer"))
pairs_pre <- list(c("Pre-SMN", "No Cancer"), c("Post-SMN", "No Cancer"), c("Pre-PMN", "No Cancer"), c("Post-PMN", "No Cancer"))


calculate_delta_cf <- function(df, pairs) {
  # Pivot data to wide format to faciliate calculation
  df_wide <- df %>%
    pivot_wider(names_from = group, values_from = norm_cum_freq, values_fill = list(norm_cum_freq = 0))

  # Ensure no NAs and consolidate for each subgroup
  df_summarized <- df_wide %>%
    group_by(frag_size, frac, variable) %>%
    summarize(across(everything(), sum, na.rm = TRUE), .groups = 'drop')

  # Calculate delta CF for each pair
  for (pair in pairs) {
    group1 <- gsub("[ -]", "_", pair[1])  # Replace spaces and hyphens with underscores
    group2 <- gsub("[ -]", "_", pair[2])
    delta_col_name <- paste0("delta_cf_", group1, "_", group2)
    
    # Calculate difference
    df_summarized <- df_summarized %>%
      mutate(!!delta_col_name := !!sym(pair[1]) - !!sym(pair[2]))
  }

  # Reshape to long format 
  df_long <- df_summarized %>%
    pivot_longer(
      cols = starts_with("delta_cf"),
      names_to = "group",
      values_to = "delta_cf"
    )

  return(df_long)
}

# Calculate delta CF
df.delta_cf.ca <- calculate_delta_cf(df.cumfreq_norm.ca, pairs_ca)
df.delta_cf.smn <- calculate_delta_cf(df.cumfreq_norm.smn, pairs_smn)
df.delta_cf.pre <- calculate_delta_cf(df.cumfreq_norm.pre, pairs_pre)

# Rename group column for figure labeling
df.delta_cf.ca$group <- recode(df.delta_cf.ca$group,
                                delta_cf_Cancer_No_Cancer = "Cancer (n = 49)")

df.delta_cf.smn$group <- recode(df.delta_cf.smn$group,
                                delta_cf_SMN_No_Cancer = "SMN (n = 26)",
                                delta_cf_PMN_No_Cancer = "PMN (n = 23)") 

df.delta_cf.pre$group <- recode(df.delta_cf.pre$group,
                                delta_cf_Pre_SMN_No_Cancer = "Pre-SMN (n = 4)", 
                                delta_cf_Pre_PMN_No_Cancer = "Pre-PMN (n = 5)",
                                delta_cf_Post_SMN_No_Cancer = "Post-SMN (n = 22)",
                                delta_cf_Post_PMN_No_Cancer = "Post-PMN (n = 18)") 

# Reorder factors for labeling
df.delta_cf.smn <- df.delta_cf.smn %>% mutate(group = factor(group, levels = c("SMN (n = 26)", "PMN (n = 23)")))
df.delta_cf.pre <- df.delta_cf.pre %>% mutate(group = factor(group, levels = c("Pre-SMN (n = 4)", "Post-SMN (n = 22)", "Pre-PMN (n = 5)", "Post-PMN (n = 18)")))


# Function to plot delta CF
plot_delta_cf <- function(df, color_palette) {
  p <- df %>%
    filter(variable != "median.freq") %>%
    group_by(frac, variable) %>%
    group_map(~ggplot(data = .x, aes(x = frag_size, y = delta_cf, color = group)) +
                geom_line(size = 0.75) +
                labs(x = "Fragment size (bp)", y = expression(paste(Delta, "CF")), title = .y[[1]]) +
                scale_color_manual(name = "Group", values = color_palette) +
                theme(axis.title.x = element_text(size = 10),
                      axis.title.y = element_text(size = 10),
                      plot.title = element_text(size = 10)))
  return(p)
}


# Define color palette
viridis_palette <- viridis(30)
show_col(viridis_palette)
color_ca <- viridis_palette[1]
color_smn <- viridis_palette[c(1, 16)]
color_pre <- viridis_palette[c(1, 9, 16, 24)] 

# Generate plot for subgroups 
p1.delta_cf.ca <- plot_delta_cf(df.delta_cf.ca, color_ca)
p1.delta_cf.smn <- plot_delta_cf(df.delta_cf.smn, color_smn)
p1.delta_cf.pre <- plot_delta_cf(df.delta_cf.pre, color_pre)

# Combine plots
p1.delta_cf.ca.combined <- ggarrange(plotlist = p1.delta_cf.ca, ncol = 3, nrow = 1, labels = c("A", "B", "C"), common.legend = TRUE, legend = "right")

p1.delta_cf.smn.combined <- ggarrange(plotlist = p1.delta_cf.smn, ncol = 3, nrow = 1, labels = c("D", "E", "F"), common.legend = TRUE, legend = "right")

p1.delta_cf.pre.combined <- ggarrange(plotlist = p1.delta_cf.pre, ncol = 3, nrow = 1, labels = c("G", "H", "I"), common.legend = TRUE, legend = "right")

# Combine all and save  
p1.delta_cf.combined <- ggarrange(p1.delta_cf.ca.combined, p1.delta_cf.smn.combined, p1.delta_cf.pre.combined, ncol = 1, nrow = 3, common.legend = F, legend = "bottom")

imagedir="protocol_ms/data/12-size_distr"
out.plot <- file.path(imagedir, "SFigure3_delta_cf_240416.tiff")
ggsave(filename = out.plot, plot = p1.delta_cf.combined, device = "tiff", dpi = 300, width = 8, height = 9, bg = "white")
 
```


