# Load Library 
```{r library, echo = FALSE}
library(tidyverse)
library(tidyr)
library(reshape2)
library(dplyr)
library(ggplot2)
library(readr)
library(ggpubr)
library(stats)
library(grid)
library(rstatix)
library(broom)
library(scales)
library(viridis)
```
# Import and Clean cfDNA Fragment/Bin Dataframe
```{r fragment_size, echo=FALSE}

#Read RDS with fragment/bin data
bins5mb <- readRDS("data/cfdna_pipeline/11-combine_bin/binfrag_summary.rds")

#load meta-data 
meta <- read_csv("data/sample_reference_sorted230325.csv")
meta <- dplyr::rename(meta, id = Legacy.Custom.ID) #rename to id


#summarise counts for each sample for each cfDNA fragment size
df.fr <- bins5mb %>%
  group_by(id) %>% 
  summarise(
    across(counts_100:counts.corrected_638, sum) #Note: columns are out of order, but includes 100-650 bp
  )

#select specific meta data and cancer type
meta <- meta %>% select(id, Cancer, pre_cancer, smn, Breast.Cancer, Thyroid.Cancer) 

#replace NA with "No"
meta <- meta %>%
  mutate_at(vars(smn), ~replace_na(., "No"))

#merge meta data with motif data
df.fr <- inner_join(meta, df.fr, by="id") 

#create phenotype subgroups
df.fr <- df.fr %>% 
  mutate(
    group = case_when(
      Cancer == "Yes" & pre_cancer == "No" ~ "Post-Dx", 
      Cancer == "Yes" & pre_cancer == "Yes" ~ "Pre-Dx",
      Cancer == "No" & pre_cancer == "No" ~ "No Cancer"), 
    group2 = case_when(
      Cancer == "Yes" & smn == "Yes" ~ "SMN",
      Cancer == "Yes" & smn == "No" ~ "PMN",
      Cancer == "No" & smn == "No" ~ "No Cancer"), 
    group3 = case_when(
      id %in% c("CCF02224-01-001", "CCF07535-01-001", "CCF02986-01-001", "CCF06617-01-001") ~ "Pre-SMN",
      id %in% c("CCF02583-01-001", "CCF02068-01-001", "CCF04317-01-001", "CCF07793-01-001", "CCF05875-02-001") ~ "Pre-PMN",
      TRUE ~ group2
    )
  ) %>% 
  mutate(
    ca_type1 = case_when(
      Cancer == "No" ~ "No Cancer",
      Breast.Cancer == "Yes" & Thyroid.Cancer == "Yes" ~ "Both",
      Breast.Cancer == "Yes" & Thyroid.Cancer == "No" ~ "Breast",
      Breast.Cancer == "No" & Thyroid.Cancer == "Yes" ~ "Thyroid"
    )
  )

# Rename levels in the group3 column 
df.fr <- df.fr %>%
  mutate(group3 = fct_recode(group3,
                             "Post-PMN" = "PMN",
                             "Post-SMN" = "SMN"))
# Relocate columns 
df.fr <- df.fr %>% 
  relocate(Cancer, .after = id) %>% 
  relocate(pre_cancer, .after = Cancer) %>%
  relocate(group, .after = pre_cancer)  %>%
  relocate(ca_type1, .after = group) %>%
  relocate(group2, .after = group) %>%
  relocate(group3, .after = group2) %>% 
  ungroup()

```

#Clean data for Fragment Distribution Analysis 
```{r cleaning, echo = FALSE}

#get uncorrected counts
df.fr2 <- df.fr %>% 
  select(matches("^id$|^Cancer$|^pre_cancer$|^group$|^group2$|^group3$|^Breast.Cancer$|^Thyroid.Cancer$|^counts_[0-9]+$"))

#reshape dataframe 
df.fr2 <- df.fr2 %>%
  pivot_longer(cols = starts_with("counts_"), 
               names_to = "frag_size",
               values_to = "counts") %>% 
  mutate(frag_size = as.numeric(str_remove(frag_size, "counts_")))

#get corrected counts 
df.fr2.corrected <- df.fr %>% 
  select(matches("^id$|^Cancer$|^pre_cancer$|^group$|^group2$|^group3$|^Breast.Cancer$|^Thyroid.Cancer$|^counts\\.corrected_[0-9]+$"))

#reshape dataframe 
df.fr2.corrected <- df.fr2.corrected %>%
  pivot_longer(cols = starts_with("counts.corrected_"), 
               names_to = "frag_size",
               values_to = "counts_corrected") %>% 
  mutate(frag_size = as.numeric(str_remove(frag_size, "counts.corrected_")))

#merge the two dataframes
df.frags <- df.fr2 %>% 
  full_join(df.fr2.corrected, 
            by = c("id", "Cancer", "pre_cancer", "group", "group2", "group3", 
                   "Breast.Cancer", "Thyroid.Cancer", "frag_size"))

#calculate freq and total counts
df.frags <- df.frags %>%
  group_by(id) %>% 
  mutate(
    total.counts = sum(counts), 
    total.counts.corrected = sum(counts_corrected),
    freq = counts / total.counts,
    freq.corrected = counts_corrected/total.counts.corrected
  ) %>% ungroup() %>%
  mutate(
    frac = case_when(
        frag_size >= 100 & frag_size <=250 ~ "mono", 
        frag_size >= 251 & frag_size <=450 ~ "di", 
        frag_size >= 451 & frag_size <=650 ~ "tri"
  )) %>% 
  relocate(frac, .before = frag_size)

#Save as csv
write.csv(df.frags, "data/cfdna_pipeline/12-fraglength_analysis/12-fraglength_summary_230323.csv", row.names = FALSE)
```

# Functions (KS test)
```{r functions, echo = FALSE}

# Define the function to perform pairwise ks tests
ks_test <- function(data, variable_name) {
  # Get the list of groups
  groups <- unique(data$group)
  
  # Generate all pairs of groups
  pairs <- combn(groups, 2, simplify = FALSE)
  
  # Perform the KS test for each pair
  results <- lapply(pairs, function(pair) {
    group1_data <- data$value[data$group == pair[1]]
    group2_data <- data$value[data$group == pair[2]]
    ks_result <- ks.test(group1_data, group2_data)
    
    # Return a data frame with the test result and the groups compared
    data.frame(
      variable = variable_name,
      group1 = pair[1],
      group2 = pair[2],
      statistic = ks_result$statistic,
      p.value = ks_result$p.value
    )
  })
  
  # Combine the results into a single data frame
  do.call(rbind, results)
}

```

# Fragment Distributions by Cancer Subgroups
```{r frag_dist, echo = FALSE}

#df for median fragment size freq by cancer status 
df.frags.ca <- df.frags %>% ungroup() %>% 
  group_by(Cancer, frag_size) %>%
  summarise(
    median.freq = median(freq),
    median.freq.correct = median(freq.corrected)
  ) %>% 
  mutate(
    frac = case_when(
        frag_size >= 100 & frag_size <=250 ~ "Mononucleosome", 
        frag_size >= 251 & frag_size <=450 ~ "Dinucleosome", 
        frag_size >= 451 & frag_size <=650 ~ "Trinucleosome"
  )) %>%
  mutate(
    frac = factor(frac, levels = c("Mononucleosome", "Dinucleosome", "Trinucleosome"))
  )


#df for median fragment size freq by SMN status 
df.frags.smn <- df.frags %>% ungroup() %>%
  group_by(group2, frag_size) %>%
  summarise(
    median.freq = median(freq),
    median.freq.correct = median(freq.corrected)
  ) %>% 
  mutate(
    frac = case_when(
        frag_size >= 100 & frag_size <=250 ~ "Mononucleosome", 
        frag_size >= 251 & frag_size <=450 ~ "Dinucleosome", 
        frag_size >= 451 & frag_size <=650 ~ "Trinucleosome"
  )) %>%
  mutate(
    frac = factor(frac, levels = c("Mononucleosome", "Dinucleosome", "Trinucleosome")), 
    group2 = factor(group2, levels = c("SMN", "PMN", "No Cancer"))
  ) 

#df for median fragment size freq by SMN Status & Plasma Draw Time 
df.frags.pre <- df.frags %>% ungroup() %>%
  group_by(group3, frag_size) %>%
  summarise(
    median.freq = median(freq),
    median.freq.correct = median(freq.corrected)
  ) %>% 
  mutate(
    frac = case_when(
        frag_size >= 100 & frag_size <=250 ~ "Mononucleosome", 
        frag_size >= 251 & frag_size <=450 ~ "Dinucleosome", 
        frag_size >= 451 & frag_size <=650 ~ "Trinucleosome"
  )) %>%
  mutate(
    frac = factor(frac, levels = c("Mononucleosome", "Dinucleosome", "Trinucleosome")), 
    group3 = factor(group3, levels = c("Pre-SMN", "Post-SMN", "Pre-PMN", "Post-PMN", "No Cancer"))
  ) 
```

# cfDNA Fragment Size Distribution for Cancer/No Cancer
```{r fig_size_ca, echo = FALSE}

# Set the sample sizes for each cancer/no cancer group
n_cancer <- 49
n_no_cancer <- 50

#set color palette 
viridis_palette <- viridis(30)
custom_color <- viridis_palette[c(1, 29)]
show_col(viridis_palette)

#plot distribution for each nucleosome fraction
p1.dist <- df.frags.ca %>%
  group_by(frac) %>% 
  group_map(~{
    p <- ggplot(.) + 
            aes(x = frag_size, y = median.freq.correct, color = Cancer) +
            scale_color_manual(name = "Group",
                                 labels = c(paste0("Cancer (N = ", n_cancer, ")"),
                                            paste0("No Cancer (N = ", n_no_cancer, ")")), values = custom_color) +
            geom_line(size = 0.75) + 
            labs(x = "Fragment size (bp)", y = "Frequency", title = .y[[1]]) + 
            theme(axis.title.x = element_text(size = 10),
                  axis.title.y = element_text(size = 10),
                  plot.title = element_text(size = 10))
    
    if (.y[[1]] == "Dinucleosome") { #added to make y-axis consistent with other figures
      p <- p + scale_y_continuous(breaks = seq(0.0004, 0.0016, by = 0.0004), 
                                  labels = function(x) ifelse(x == 0, '', format(x, scientific = FALSE)),
                                  limits = c(0, 0.001625),  # extend the y-axis limits
                                  expand = c(0,0))  # remove default expansion
    }
    
    p
  })



#arrange plots in grid
p1.dist.ca.combined <- ggarrange(p1.dist[[1]], p1.dist[[2]], p1.dist[[3]], 
                             ncol = 3, nrow = 1, labels = c("A", "B", "C"), common.legend = TRUE, legend = "right")

```

# cfDNA Fragment Size Distribution by SMN Status 
```{r fig_size_ca, echo = FALSE}

# Set the sample sizes for each group
n_smn <- 26
n_pmn <- 23
n_no_cancer <- 50

# Update group names with sample sizes
df.frags.smn <- df.frags.smn %>%
  mutate(group2 = recode(group2,
                         `SMN` = paste0("SMN (N = ", n_smn, ")"),
                         `PMN` = paste0("PMN (N = ", n_pmn, ")"),
                         `No Cancer` = paste0("No Cancer (N = ", n_no_cancer, ")")))

#pick colors
viridis_palette <- viridis(30)
custom_color <- viridis_palette[c(1, 16, 29)]
show_col(viridis_palette)

#plot distribution for each nucleosome fraction
p1.dist.smn <- df.frags.smn %>%
  group_by(frac) %>% 
  group_map(~ggplot(.) + 
            aes(x = frag_size, y = median.freq.correct, color = group2) +
            geom_line(size = 0.75) + 
            labs(x = "Fragment size (bp)", y = "Frequency", title = .y[[1]]) +
            scale_color_manual(name = "Group",
                               values = custom_color) +
            theme(axis.title.x = element_text(size = 10),
                  axis.title.y = element_text(size = 10),
                  plot.title = element_text(size = 10)))

#combine plots 
p1.dist.smn.combined <- ggarrange(plotlist = p1.dist.smn, 
                             ncol = 3, nrow = 1, labels = c("D", "E", "F"), common.legend = TRUE, legend = "right")

```

# cfDNA Fragment Size Distribution by SMN Status & Plasma Draw Time 
```{r fig_size_ca, echo = FALSE}

# Set the sample sizes for each group
n_smn <- 4
n_postsmn <- 22
n_pmn <- 5
n_postpmn <- 18
n_no_cancer <- 50

# Update group names with sample sizes
df.frags.pre <- df.frags.pre %>%
  mutate(group3 = recode(group3,
                         `Pre-SMN` = paste0("Pre-SMN (N = ", n_smn, ")"),
                         `Pre-PMN` = paste0("Pre-PMN (N = ", n_pmn, ")"),
                         `Post-SMN` = paste0("Post-SMN (N = ", n_postsmn, ")"),
                         `Post-PMN` = paste0("Post-PMN (N = ", n_postpmn, ")"),
                         `No Cancer` = paste0("No Cancer (N = ", n_no_cancer, ")")))

#Set color palette 
viridis_palette <- viridis(30)
custom_color <- viridis_palette[c(1, 9, 16, 24, 29)]
show_col(viridis_palette)


#plot distribution for each nucleosome fraction with modified line width and color
p1.dist.pre <- df.frags.pre %>% 
  group_by(frac) %>% 
  group_map(~{
    group3_numeric <- as.numeric(factor(.$group3))
    ggplot(.) + 
      aes(x = frag_size, y = median.freq.correct, color = group3) +
      geom_line(size = 0.75) + 
      labs(x = "Fragment size (bp)", y = "Frequency",  title = .y[[1]]) +
      scale_color_manual(values = custom_color, name = "Group") +
      guides(linetype = FALSE, size = FALSE) + 
      theme(
        axis.title.x = element_text(size = 10),
        axis.title.y = element_text(size = 10),
        plot.title = element_text(size = 10))
  })

#arrange plots in grid
p1.dist.pre.combined <- ggarrange(plotlist = p1.dist.pre, ncol = 3, nrow = 1, labels = c("G", "H", "I"), common.legend = TRUE, legend = "right")


#combine all distirbution plots
p1.dist.combined <- ggarrange(p1.dist.ca.combined, p1.dist.smn.combined, p1.dist.pre.combined, ncol = 1, nrow = 3, common.legend = F, legend = "right")

#create file path for figure
out.plot <- file.path(imagedir, "Figure1_fragdist_combined.tiff")

#save as tiff
ggsave(filename = out.plot, plot = p1.dist.combined, device = "tiff", dpi = 300, width = 8, height = 9)
```

# Perform KS test - All Nucleomes
```{r ks_test_all, echo = FALSE}

#reshape data for cancer/vs no cancer
df_long.ca <- df.frags.ca %>%
  select(Cancer, median.freq.correct, median.freq, frac) %>%
  pivot_longer(cols = c("median.freq.correct", "median.freq"), names_to = "variable", values_to = "value") %>%
  rename(group = Cancer)
 
 
# Group the data by the variable and split it into a list of data frames
df_grouped.ca <- split(df_long.ca, df_long.ca$variable)

# Apply the ks test to each group
results.ca <- lapply(names(df_grouped.ca), function(name) ks_test(df_grouped.ca[[name]], name))


# Combine the results into a single data frame
df_results.ca <- do.call(rbind, results.ca)

#SMN status
df_long.smn <- df.frags.smn %>%
  select(group2, median.freq.correct, median.freq, frac) %>%
  pivot_longer(cols = c("median.freq.correct", "median.freq"), names_to = "variable", values_to = "value") %>%
  rename(group = group2)


# Group the data by the variable and split it into a list of data frames
df_grouped.smn <- split(df_long.smn, df_long.smn$variable)

# Apply the ks test to each group
results.smn <- lapply(names(df_grouped.smn), function(name) ks_test(df_grouped.smn[[name]], name))

# Combine the results into a single data frame
df_results.smn <- do.call(rbind, results.smn)


#Pre-Malignant status
df_long.pre <- df.frags.pre %>%
  select(group3, median.freq.correct, median.freq, frac) %>%
  pivot_longer(cols = c("median.freq.correct", "median.freq"), names_to = "variable", values_to = "value")%>%
  rename(group = group3)

# Group the data by the variable and split it into a list of data frames
df_grouped.pre <- split(df_long.pre, df_long.smn$variable)

# Apply the ks test to each group
results.pre <- lapply(names(df_grouped.pre), function(name) ks_test(df_grouped.pre[[name]], name))


# Combine the results into a single data frame
df_results.pre <- do.call(rbind, results.pre)

#merge all
df_results.merg <- rbind(df_results.ca, df_results.smn, df_results.pre)

#remove rownames
rownames(df_results.merg) <- NULL

#merge and relocate
df_results.merg <- df_results.merg %>% relocate(variable, .after = group2)

#save to csv 
write.csv(df_results.merg, "data/cfdna_pipeline/12-fraglength_analysis/12-kstest_overall_pairwise.csv", row.names = FALSE)

```

# Perform KS Test - Individual nucleosome fraction 
```{r ks_test_frac, echo = FALSE}

# Load the necessary library
library(stringr)

## Cancer status
# Group the data by the variable and frac and split it into a list of data frames
df_grouped.ca <- split(df_long.ca, list(df_long.ca$variable, df_long.ca$frac))

# Apply the ks test to each group, passing the name of the variable and frac as an argument
results.ca <- lapply(names(df_grouped.ca), function(name) ks_test(df_grouped.ca[[name]], name))

# Combine the results into a single data frame and remove row names
df_results.ca <- do.call(rbind, results.ca)

# Split the 'variable' column
df_results.ca <- df_results.ca %>%
  mutate(frac = str_extract(variable, "\\.[^.]+$")) %>%  # Extract the string after the last dot
  mutate(frac = str_remove(frac, "\\.")) %>%  # Remove the dot from the beginning of the frac
  mutate(variable = str_remove(variable, "\\.[^.]+$")) %>%# Remove the frac part from the variable 
  relocate(variable, .after = group2) %>%
  relocate(frac, .after = variable)

##SMN status
# Group the data by the variable and frac and split it into a list of data frames
df_grouped.smn <- split(df_long.smn, list(df_long.smn$variable, df_long.smn$frac))

# Apply the ks test to each group, passing the name of the variable and frac as an argument
results.smn <- lapply(names(df_grouped.smn), function(name) ks_test(df_grouped.smn[[name]], name))

# Combine the results into a single data frame and remove row names
df_results.smn <- do.call(rbind, results.smn)

# Split the 'variable' column
df_results.smn <- df_results.smn %>%
  mutate(frac = str_extract(variable, "\\.[^.]+$")) %>%  # Extract the string after the last dot
  mutate(frac = str_remove(frac, "\\.")) %>%  # Remove the dot from the beginning of the frac
  mutate(variable = str_remove(variable, "\\.[^.]+$")) %>%# Remove the frac part from the variable 
  relocate(variable, .after = group2) %>%
  relocate(frac, .after = variable)

## SMN Status & Plasma Draw Time 
# Group the data by the variable and frac and split it into a list of data frames
df_grouped.pre <- split(df_long.pre, list(df_long.pre$variable, df_long.pre$frac))

# Apply the ks test to each group, passing the name of the variable and frac as an argument
results.pre <- lapply(names(df_grouped.pre), function(name) ks_test(df_grouped.pre[[name]], name))

# Combine the results into a single data frame and remove row names
df_results.pre <- do.call(rbind, results.pre)

# Split the 'variable' column
df_results.pre <- df_results.pre %>%
  mutate(frac = str_extract(variable, "\\.[^.]+$")) %>%  # Extract the string after the last dot
  mutate(frac = str_remove(frac, "\\.")) %>%  # Remove the dot from the beginning of the frac
  mutate(variable = str_remove(variable, "\\.[^.]+$")) %>%# Remove the frac part from the variable 
  relocate(variable, .after = group2) %>%
  relocate(frac, .after = variable)
  
#merge all
df_results.merge <- rbind(df_results.ca, df_results.smn, df_results.pre)

rownames(df_results.merge) <- NULL

#save to csv 
write.csv(df_results.merge, "data/cfdna_pipeline/12-fraglength_analysis/12-kstest_pernuc_pairwise.csv", row.names = FALSE)

```
# Summarize Major Peaks by Cancer Subgroups
```{r summarize_majorpeak, echo = FALSE}

# Reshape the dataframe
df.frags_long <- df.frags %>%
  pivot_longer(cols = c(freq, freq.corrected),
               names_to = "type",
               values_to = "value") %>%
  select(id, Cancer, group2, group3, frag_size, frac, type, value) %>%
  mutate(
    frac = factor(frac, levels = c("mono", "di", "tri"))
  )

#get max fragment size and q25 and q75
df.frags_summary <- df.frags_long %>%
  group_by(id, Cancer, group2, group3, frac, type) %>%
  summarise(max_frag_size = frag_size[which.max(value)],
            q25_frag_size = quantile(frag_size, 0.25, na.rm = TRUE),
            q75_frag_size = quantile(frag_size, 0.75, na.rm = TRUE)) %>%
  ungroup()
 

#summarize max fragment size for Cancer/No Cancer
df.peak.ca <- df.frags_summary %>%
  group_by(Cancer, frac, type) %>%
  summarise(
    max_peak = median(max_frag_size),
    q25 = quantile(max_frag_size, 0.25), 
    q75 = quantile(max_frag_size, 0.75)
  ) %>%
  rename(group = Cancer)

#summarize max fragment size by SMN Status
df.peak.smn <- df.frags_summary %>%
  group_by(group2, frac, type) %>%
  summarise(
    max_peak = median(max_frag_size),
    q25 = quantile(max_frag_size, 0.25), 
    q75 = quantile(max_frag_size, 0.75)
  ) %>%
  rename(group = group2)

#summarize max fragment size by SMN Status & Plasma Draw Time
df.peak.pre <- df.frags_summary %>%
  group_by(group3, frac, type) %>%
  summarise(
    max_peak = median(max_frag_size),
    q25 = quantile(max_frag_size, 0.25), 
    q75 = quantile(max_frag_size, 0.75)
  ) %>% 
  rename(group = group3)

#merge
df.peak.merge <- rbind(df.peak.ca, df.peak.smn, df.peak.pre)
```

# Statistical Test Comparing Major cfDNA peaks 
```{r compare_majorpeaks, echo = FALSE}
#KW test for overall group testing by SMN status
overall.max.smn <- df.frags_summary %>%
  group_by(frac, type) %>%
  kruskal_test(max_frag_size ~ group2)

#KW test for overall group testing by SMN Status & Plasma Draw Time 
overall.max.pre <- df.frags_summary %>%
  group_by(frac, type) %>%
  kruskal_test(max_frag_size ~ group3)

#merge overall results
overall.merge <- rbind(overall.max.smn, overall.max.pre)

#write to csv
write.csv(overall.merge, "data/cfdna_pipeline/12-fraglength_analysis/12-maxpeak_overall_kwtest.csv", row.names =F) 


#pairwise Wilcox Rank Sum Test by cancer/No Cancer
peak.test <- df.frags_summary %>%
  group_by(frac, type)%>%
  wilcox_test(max_frag_size ~ Cancer) %>%
  add_significance() %>%
  select(-.y.)

#pairwise Wilcox Rank Sum Test by SMN Status
peak.test2<- df.frags_summary %>%
  group_by(frac, type)%>%
  wilcox_test(max_frag_size ~ group2) %>%
  add_significance() %>%
  select(-.y., -p.adj) %>%
  rename(p.signif = p.adj.signif)

#pairwise Wilcox Rank Sum Test by SMN Status & Plasma Draw Time 
peak.test3<- df.frags_summary %>%
  group_by(frac, type)%>%
  wilcox_test(max_frag_size ~ group3) %>%
  add_significance() %>%
  select(-.y., -p.adj) %>%
  rename(p.signif = p.adj.signif)

#merge all for supplelmentary table 
peak.merge <- rbind(peak.test, peak.test2, peak.test3)

#write to csv
write.csv(peak.merge, "data/cfdna_pipeline/12-fraglength_analysis/12-maxpeak_wilcox.csv", row.names =F) 


```
