---
title: "cfDNA fragment end motif"
author: "Darren Liu"
date: "2023-12-13"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
#For PC
knitr::opts_knit$set(root.dir = "W:/Shared/LRI/Labs/engclab/LAB_MEMBER_CENTRAL/Darren_Liu/3_cf-DNA")


```

# Load library
```{r libraries}
library(tidyverse)
library(ggpubr)
library(rstatix)
library(stringr)
library(RColorBrewer)
library(scales)
library(viridis)
library(pheatmap)
library(colorspace) #for colorblind friendly colors
library(gridExtra)
library(grid)
```

# Import and Clean End Motif RDS file
```{r import_clean}

#import end nmotif rds file 
df.motif <- readRDS("data/cfdna_pipeline/21-combine_motif_4mer/endmotif_4bp_gc_summary.rds")

#summarize counts as dataframe still stratified by gc-strata
df.motif <- df.motif %>%
  group_by(id, end_motif, frac) %>%
  summarise(
    count = sum(count), 
    count.corrected = sum(count.corrected), 
    frac.total = first(frac.total),
    frac.total.corrected = first(frac.total.corrected)
  )
  
#Import meta data
meta <- read_csv("data/sample_reference_sorted230325.csv")
meta <- dplyr::rename(meta, id = Legacy.Custom.ID) #rename to id

#select specific meta data and cancer type
meta <- meta %>% select(id, Cancer, pre_cancer, smn, Breast.Cancer, Thyroid.Cancer, ca_1st, ca_2nd, ca_3rd, ca_4th) 

#replace NA with "No"
meta <- meta %>%
  mutate_at(vars(smn), ~replace_na(., "No")) %>%
  mutate_at(vars(ca_1st, ca_2nd, ca_3rd, ca_4th), ~replace_na(., "None"))

#merge meta data with motif data
summary.motif <- inner_join(meta, df.motif, by="id") 

#create phenotype subgroups
summary.motif <- summary.motif %>% 
  mutate(
    group = case_when(
      Cancer == "Yes" & pre_cancer == "No" ~ "Post-Dx", 
      Cancer == "Yes" & pre_cancer == "Yes" ~ "Pre-Dx",
      Cancer == "No" & pre_cancer == "No" ~ "No Cancer"), 
    group2 = case_when(
      Cancer == "Yes" & smn == "Yes" ~ "SMN",
      Cancer == "Yes" & smn == "No" ~ "PMN",
      Cancer == "No" & smn == "No" ~ "No Cancer"), 
    group3 = case_when(
      id %in% c("CCF02224-01-001", "CCF07535-01-001", "CCF02986-01-001", "CCF06617-01-001") ~ "Pre-SMN",
      id %in% c("CCF02583-01-001", "CCF02068-01-001", "CCF04317-01-001", "CCF07793-01-001", "CCF05875-02-001") ~ "Pre-PMN",
      TRUE ~ group2
    )
  ) %>% 
  mutate(
    ca_type1 = case_when(
      Cancer == "No" ~ "No Cancer",
      Breast.Cancer == "Yes" & Thyroid.Cancer == "Yes" ~ "Both",
      Breast.Cancer == "Yes" & Thyroid.Cancer == "No" ~ "Breast",
      Breast.Cancer == "No" & Thyroid.Cancer == "Yes" ~ "Thyroid"
    )
  )

# Rename levels in the group3 column 
summary.motif <- summary.motif %>%
  mutate(group3 = fct_recode(group3,
                             "Post-PMN" = "PMN",
                             "Post-SMN" = "SMN"))

# Relocate columns 
summary.motif <- summary.motif %>% 
  relocate(Cancer, .after = id) %>% 
  relocate(pre_cancer, .after = Cancer) %>%
  relocate(group, .after = pre_cancer)  %>%
  relocate(ca_type1, .after = group) %>%
  relocate(group2, .after = group) %>%
  relocate(group3, .after = group2) %>% 
  ungroup()

# #Save as csv
write.csv(summary.motif, "data/cfdna_pipeline/22-summarize_motif_4mer/22-summary.motif.4bp.csv", row.names = FALSE)
```

# Summary Statistics, Related to Figure 4A    
```{r summary_stats, echo = FALSE}

#End Motif Profile by SMN status
summary.motif.smn <- summary.motif %>% 
  group_by(id, end_motif, group2) %>%
  summarise(
    count = sum(count), 
    count.corrected = sum(count.corrected)
  ) %>% ungroup() %>% 
  group_by(id) %>% 
  mutate(
    total = sum(count),
    total.corrected = sum(count.corrected),
    freq = count / total, 
    freq.corrected = count.corrected / total.corrected
  ) %>%
  mutate(
    group2 = factor(group2, levels = c("SMN", "PMN", "No Cancer"))
  )

#End Motif Profile by SMN Status and Draw Time 
summary.motif.pre <- summary.motif %>% 
  group_by(id, end_motif, group3) %>%
  summarise(
    count = sum(count), 
    count.corrected = sum(count.corrected)
  ) %>% ungroup() %>% 
  group_by(id) %>% 
  mutate(
    total = sum(count),
    total.corrected = sum(count.corrected),
    freq = count / total, 
    freq.corrected = count.corrected / total.corrected
  ) %>%
  mutate(group3 = factor(group3, levels = c("Pre-SMN", "Post-SMN", "Pre-PMN", "Post-PMN", "No Cancer")))

# Overall testing of 4-mer by SMN status
stat.smn.overall <- summary.motif.smn %>%
  group_by(end_motif) %>%
  summarise(p.value = kruskal.test(freq.corrected ~ group2)$p.value)

# Filter for significant only 
smn.sig.4mer <- stat.smn.overall %>% filter(p.value < 0.05) #%>% select(end_motif)

# Pairwise comparison of statistically significant 
# Wilcox test by smn status (for only significant 4-mer)
stat.smn <- summary.motif.smn %>%
  filter(end_motif %in% smn.sig.4mer$end_motif) %>% 
  group_by(end_motif) %>%
  wilcox_test(freq.corrected ~ group2) %>%
  adjust_pvalue(method = "BH") %>%
  add_significance() %>%
  select(-.y.)%>%
  arrange(p)

# write.csv(stat.smn.overall, "data/cfdna_pipeline/22-summarize_motif_4mer/22-4mer.smn.kwtest.csv", row.names = FALSE)
# write.csv(smn.sig.4mer, "data/cfdna_pipeline/22-summarize_motif_4mer/22-4mer.smn.kwtest.SIG.csv", row.names = FALSE)
# write.csv(stat.smn, "data/cfdna_pipeline/22-summarize_motif_4mer/22-4mer.smn.wilcox.SIG.csv", row.names = FALSE)

# Overall testing of 4-mer by SMN status and plasma draw time 
stat.pre.overall <- summary.motif.pre %>%
  group_by(end_motif) %>%
  summarise(p.value = kruskal.test(freq.corrected ~ group3)$p.value)

# Filter for significant 4-mers
pre.sig.4mer <- stat.pre.overall %>% filter(p.value < 0.05) #%>% select(end_motif)


# Pairwise comparison of statistically significant 
# Wilcox test by SMN status/plasma draw time (for only significant 4-mer)
stat.pre <- summary.motif.pre %>%
  filter(end_motif %in% pre.sig.4mer$end_motif) %>%
  group_by(end_motif) %>%
  wilcox_test(freq.corrected ~ group3) %>%
  adjust_pvalue(method = "BH") %>%
  add_significance() %>%
  select(-.y.) 


# write.csv(stat.pre.overall, "data/cfdna_pipeline/22-summarize_motif_4mer/22-4mer.pre.kwtest.csv", row.names = FALSE)
# write.csv(pre.sig.4mer, "data/cfdna_pipeline/22-summarize_motif_4mer/22-4mer.pre.kwtest.SIG.csv", row.names = FALSE)
#write.csv(stat.pre, "data/cfdna_pipeline/22-summarize_motif_4mer/22-4mer.pre.wilcox.SIG.allcsv", row.names = FALSE)

# Combine the significant 4-mer from both stratification groups 
combined_df <- rbind(pre.sig.4mer, smn.sig.4mer)

# Remove duplicates based on the end_motif column (will be used for heatmap)
unique_df <- combined_df %>%
  distinct(end_motif, .keep_all = TRUE)
```

# Figure 4A, Heatmap of Significantly Differentially Abundant 4-mers 
```{r summary, echo = FALSE}

#Read in summarized 4-mer end motif 
summary.motif <- read.csv("data/cfdna_pipeline/22-summarize_motif_4mer/22-summary.motif.4bp.csv")


# End Motif Profile by ID
df.motif <- summary.motif %>% 
  group_by(id, end_motif, Cancer, pre_cancer, group, group2, group3, ca_type1, ca_1st, ca_2nd, ca_3rd, ca_4th) %>% 
  summarise(
    count = sum(count), 
    count.corrected = sum(count.corrected)
  ) %>% ungroup() %>% 
  group_by(id) %>% 
  mutate(
    total = sum(count),
    total.corrected = sum(count.corrected),
    freq = count / total, 
    freq.corrected = count.corrected / total.corrected
  ) %>% ungroup() 

# Only significant 4-mers
df.motif <- df.motif %>% filter(end_motif %in% unique_df$end_motif)

# Pivot the data to wide format
df.motif_wide <- df.motif %>%
  select(id, end_motif, freq.corrected) %>%
  pivot_wider(names_from = id, values_from = freq.corrected)

# Convert to matrix 
heatmap_matrix <- as.matrix(df.motif_wide[,-1])
rownames(heatmap_matrix) <- df.motif_wide$end_motif

# Create annotation dataframe for columns
annotation_data <- df.motif %>%
  select(id, group2, group3, ca_1st, ca_2nd) %>%
  distinct() %>%
  arrange(id) %>%
  column_to_rownames("id")

# Using the Paired palette
#colors <- brewer.pal(8, "Paired")

viridis_palette <- viridis(30)
colors <- viridis_palette[c(1, 9, 16, 24, 29)]

# Mapping the colors
group2_colors <- c("SMN" = colors[1], "PMN" = colors[3], "No Cancer" = colors[5])
group3_colors <- c("Pre-SMN" = colors[1], "Post-SMN" = colors[2], 
                   "Pre-PMN" = colors[3], "Post-PMN" = colors[4], 
                   "No Cancer" = colors[5])


# Identify all unique cancer types
all_cancers <- unique(c(annotation_data$ca_1st, annotation_data$ca_2nd))
cancer_colors <- brewer.pal(length(all_cancers), "Dark2")

# Create a named vector to map cancer types to colors
color_mapping <- setNames(cancer_colors, all_cancers)


# Rename columns in the annotation data frame
names(annotation_data)[names(annotation_data) == "group2"] <- "SMN"
names(annotation_data)[names(annotation_data) == "group3"] <- "Draw Time"
names(annotation_data)[names(annotation_data) == "ca_1st"] <- "1st Ca."
names(annotation_data)[names(annotation_data) == "ca_2nd"] <- "2nd Ca."


annotation_colors <- list(
    SMN = group2_colors,         # Changed 'group2' to 'SMN'
    `Draw Time` = group3_colors, # Changed 'group3' to 'Draw Time'
    `1st Ca.` = color_mapping,   # Changed 'ca_1st' to '1st Ca.'
    `2nd Ca.` = color_mapping    # Changed 'ca_2nd' to '2nd Ca.'
)

# Generate heatmap
p1 <- pheatmap(heatmap_matrix,
         scale = "row",
         cluster_rows = TRUE, 
         cluster_cols = TRUE, 
         show_rownames = TRUE, # Assuming you might have many motifs, otherwise set to TRUE if you want to see them
         show_colnames = FALSE, 
         border_color = NA, 
         treeheight_row = 0, 
         #treeheight_col = 0,
         color = viridis::viridis(120, option = "magma"),
         height = 8,
         width = 8, 
         annotation_col = annotation_data,
         annotation_colors = annotation_colors) # Depending on the number of samples, adjust this as necessary


#save file
imagedir="accepted_manuscript"

#create file path
out.plot <- file.path(imagedir, "Fig3_endmotif4bp.tiff")

#save as tiff
ggsave(filename = out.plot, plot = p1, device = "tiff", dpi = 300, width = 6, height = 8)


```

# Import and Summarise 2-mer End Motif Data
```{r create_df, echo = FALSE}

summary.motif <- read.csv("data/cfdna_pipeline/22-summarize_motif/22-endmotif_summary.csv")

#End Motif Profile by SMN status
summary.motif.smn <- summary.motif %>% 
  group_by(id, end_motif, group2) %>%
  summarise(
    count = sum(count), 
    count.corrected = sum(count.corrected)
  ) %>% ungroup() %>% 
  group_by(id) %>% 
  mutate(
    total = sum(count),
    total.corrected = sum(count.corrected),
    freq = count / total, 
    freq.corrected = count.corrected / total.corrected
  ) %>%
  mutate(
    group2 = factor(group2, levels = c("SMN", "PMN", "No Cancer"))
  )

# Set the sample sizes for each group
n_smn <- 26
n_pmn <- 23
n_no_cancer <- 50

# Update group names with sample sizes
summary.motif.smn <- summary.motif.smn %>%
  mutate(group2 = recode(group2,
                         `SMN` = paste0("SMN (N = ", n_smn, ")"),
                         `PMN` = paste0("PMN (N = ", n_pmn, ")"),
                         `No Cancer` = paste0("No Cancer (N = ", n_no_cancer, ")")))

# Reshape dataframe using pivot_longer (for boxplot)
summary.motif.smn.long <- summary.motif.smn %>%
  pivot_longer(cols = c(freq, freq.corrected),
               names_to = "type",
               values_to = "value")


# End Motif Profile by SMN Status and Draw Time 
summary.motif.pre <- summary.motif %>% 
  group_by(id, end_motif, group3) %>%
  summarise(
    count = sum(count), 
    count.corrected = sum(count.corrected)
  ) %>% ungroup() %>% 
  group_by(id) %>% 
  mutate(
    total = sum(count),
    total.corrected = sum(count.corrected),
    freq = count / total, 
    freq.corrected = count.corrected / total.corrected
  ) %>%
  mutate(group3 = factor(group3, levels = c("Pre-SMN", "Post-SMN", "Pre-PMN", "Post-PMN", "No Cancer")))


# Reshape 
summary.motif.pre.long <- summary.motif.pre %>%
  pivot_longer(cols = contains("freq"), 
               names_to = "type", 
               values_to = "value") 

# Set the sample sizes for each group
n_smn <- 4
n_postsmn <- 22
n_pmn <- 5
n_postpmn <- 18
n_no_cancer <- 50

# Update group names with sample sizes
summary.motif.pre <- summary.motif.pre %>%
  mutate(group3 = recode(group3,
                         `Pre-SMN` = paste0("Pre-SMN (N = ", n_smn, ")"),
                         `Pre-PMN` = paste0("Pre-PMN (N = ", n_pmn, ")"),
                         `Post-SMN` = paste0("Post-SMN (N = ", n_postsmn, ")"),
                         `Post-PMN` = paste0("Post-PMN (N = ", n_postpmn, ")"),
                         `No Cancer` = paste0("No Cancer (N = ", n_no_cancer, ")")))
```

# Calculate 1-mer from 2-mer Data by SMN Status
```{r smn_1bp, echo = FALSE}

# Summarize 1-mer end motifs by SMN status 
summary.motif.smn.mono <- summary.motif.smn %>%
  #obtain mono-nucleotide motifs 
  mutate(mono_motif = str_extract(end_motif, "\\b[A-Z]{1}")) %>%
  group_by(id, mono_motif, group2) %>%
  summarise(
    freq = sum(freq),
    freq.corrected = sum(freq.corrected)
  ) %>%
  mutate(
    group2 = factor(group2, levels = c("SMN (N = 26)", "PMN (N = 23)", "No Cancer (N = 50)"))
  )


# Reshapelong 
summary.motif.frac.long <- summary.motif.smn.mono %>%
  pivot_longer(cols = contains("freq"), 
               names_to = "type", 
               values_to = "value") #get only GC corrected counts

# Rehshape wide
summary.motif.smn.mono_wide <- summary.motif.smn.mono %>%
  pivot_wider(
    id_cols = id,
    names_from = mono_motif,
    values_from = c(freq, freq.corrected),
    names_glue = "{mono_motif}.{.value}"
  )


# Overall testing 
overall.mono.smn <- summary.motif.frac.long %>%
  group_by(mono_motif, type) %>%
  kruskal_test(value ~ group2)

# Parwise comparison  
stat.test.1bp.smn <- summary.motif.frac.long %>%
  group_by(mono_motif, type) %>%
  wilcox_test(value ~ group2) %>%
  adjust_pvalue(method = "none") %>%
  add_significance() %>%
  select(-.y.) 


#save summary as csv
# write.csv(summary.motif.smn.mono_wide, "data/cfdna_pipeline/22-summarize_motif/22-summary.endmotif.1bp.csv", row.names =F) 
# write.csv(overall.mono.smn, "data/cfdna_pipeline/22-summarize_motif/22-summary.endmotif.smn.1bp.kwtest.csv", row.names =F) 
# write.csv(stat.test.1bp.smn, "data/cfdna_pipeline/22-summarize_motif/22-summary.endmotif.smn.1bp.wilcox.csv", row.names =F) 


#pick colors
viridis_palette <- viridis(30)
custom_color <- viridis_palette[c(1, 16, 29)]
show_col(viridis_palette)

#my_comparisons <- list(c("SMN", "No Cancer"), c("SMN", "PMN"), c("PMN", "No Cancer"))

#only significant comparioson
my_comparisons <- list(c("SMN (N = 26)", "No Cancer (N = 50)"), c("PMN (N = 23)", "No Cancer (N = 50)"))


#boxplot
bp1.mono.smn <- ggplot(summary.motif.smn.mono, aes(x = group2, y = freq.corrected, fill = group2)) + 
  geom_boxplot() +
  #scale_fill_brewer(palette="Pastel2", name = "Cancer Type") +
  #scale_fill_discrete(name = "Group") + 
  scale_fill_manual(name = "Group", values = custom_color) + # Use custom colors
  facet_grid(. ~ mono_motif) + 
  labs(x = "Group", y = "End Motif Frequency") +
  theme(axis.text.x = element_blank()) +
  # stat_compare_means(aes(label = paste0("p=", ..p.format..)),
  #                    label.x = 1.75, label.y = 0.4, size = 3) + #global p value
  stat_compare_means(comparisons = my_comparisons, aes(label = ..p.format..), size = 2.5, label.y =c(0.37, 0.355, 0.345))

#combine plots
p.smn.motif.combined <- ggarrange(bp, bp1.mono.smn, ncol = 1, nrow = 2, labels = c("B", "C"), common.legend = TRUE, legend = "none")


#write for-loop to save files 
imagedir="data/cfdna_pipeline/22-summarize_motif"


out.plot <- file.path(imagedir, "endmotif_smn_monodi_combined.tiff")

#save as tiff
ggsave(filename = out.plot, plot = p.smn.motif.combined, device = "tiff", dpi = 300, width = 8, height = 6)



#combine both
p.combined <- ggarrange(p.smn.motif.combined, bp.pre.combined, ncol = 1, nrow = 2, common.legend = FALSE, legend = "none")

```

# Calculate 1-mer from 2-mder Data by SMN Status & Plasma Draw Time
```{r pre_1bp, echo = FALSE}

# Calculate 1-mer from 2-mer df 
summary.motif.mono.pre <- summary.motif.pre %>%
 mutate(mono_motif = str_extract(end_motif, "\\b[A-Z]{1}")) %>%
  group_by(id, mono_motif, group3) %>%
  summarise(
    mono_sum = sum(freq),
    mono_sum.corrected = sum(freq.corrected)
  )

# Reshape long
summary.motif.mono.pre.long <- summary.motif.mono.pre %>%
  pivot_longer(cols = contains("mono_sum"), 
               names_to = "type", 
               values_to = "value") 


# Overall testing
overall.mono.pre <- summary.motif.mono.pre.long %>%
  group_by(mono_motif, type) %>%
  kruskal_test(value ~ group3)


# Pairwise testing
stat.test.1bp.pre <- summary.motif.mono.pre.long %>%
  group_by(mono_motif, type) %>%
  wilcox_test(value ~ group3) %>%
  adjust_pvalue(method = "none") %>%
  add_significance() %>%
  select(-.y.) 


```

# Box plot for 2-mer End Motif by SMN Status 
```{r bp_dismn, echo = FALSE}

#Define comparison groups 
my_comparisons <- list(c("SMN (N = 26)", "No Cancer (N = 50)"), c("PMN (N = 23)", "No Cancer (N = 50)"))

#Set color palette 
viridis_palette <- viridis(30)
custom_color <- viridis_palette[c(1, 16, 29)]
show_col(viridis_palette)


#Create box plot facetted by end-motif
bp.smn <- ggplot(summary.motif.smn, aes(x = group2, y = freq.corrected, fill = group2)) + 
  geom_boxplot() +
  facet_grid(. ~ end_motif) + 
  scale_fill_manual(name = "Group", values = custom_color) + # Use custom colors
  labs(x = "Group", y = "End Motif Frequency") +
  theme(axis.text.x = element_blank()) +
  #stat_compare_means(aes(label = paste0("P=", ..p.format..)), label.y = 0.155, size = 2) + #add global p
  stat_compare_means(comparisons = my_comparisons, aes(label = ..p.signif..), label.y = c(0.15, 0.143), size = 2.5)

# #Define image directory 
# imagedir="data/cfdna_pipeline/22-summarize_motif"
# 
# #Create file path
# out.plot <- file.path(imagedir, "endmotif.2bp.smn.tiff")
# 
# #Save as tiff
# ggsave(filename = out.plot, plot = bp, device = "tiff", dpi = 300, width = 8, height = 3)

```

# Box Plot for 1-mer  End Motif by SMN Status
```{r bp_monosmn, echo = FALSE}

#Define comparison groups
my_comparisons <- list(c("SMN (N = 26)", "No Cancer (N = 50)"), c("SMN (N = 26)", "PMN (N = 23)"), c("PMN (N = 23)", "No Cancer (N = 50)"))

#Set color palette 
viridis_palette <- viridis(30)
custom_color <- viridis_palette[c(1, 16, 29)]
show_col(viridis_palette)


#Create box plot 
bp.mono.smn <- ggplot(summary.motif.smn.mono, aes(x = group2, y = freq.corrected, fill = group2)) + 
  geom_boxplot() +
  scale_fill_manual(name = "Group", values = custom_color) + # Use custom colors
  facet_grid(. ~ mono_motif) + 
  labs(x = "Group", y = "End Motif Frequency") +
  theme(axis.text.x = element_blank()) +
  # stat_compare_means(aes(label = paste0("p=", ..p.format..)),
  #                    label.x = 1.75, label.y = 0.4, size = 3) + #global p value
  stat_compare_means(comparisons = my_comparisons, aes(label = ..p.format..), size = 2.5, label.y =c(0.37, 0.355, 0.345))

#Combine mono- and di-nucleotide end motif plots together 
p.smn.motif.combined <- ggarrange(bp.mono.smn, bp.smn, ncol = 1, nrow = 2, labels = c("B", "C"), common.legend = TRUE, legend = "bottom")



```

# Box Plot for 2-mer by SMN Status & Plasma Draw Time
```{r bp_dipre, echo = FALSE}

#Define comparison groups
my_comparisons <- list(c("Pre-SMN (N = 4)", "No Cancer (N = 50)"), 
                       c("Pre-SMN (N = 4)", "Post-SMN (N = 22)"))

#Set color palette 
viridis_palette <- viridis(30)
custom_color <- viridis_palette[c(1, 9, 16, 24, 29)]
show_col(viridis_palette)

#Create box plot facetted by end-motif
bp.2bp.pre <- ggplot(summary.motif.pre, aes(x = group3, y = freq.corrected, fill = group3)) + 
  geom_boxplot() +
  facet_grid(. ~ end_motif) + 
  scale_fill_manual(name = "Group", values = custom_color)+
  labs(x = "Group", y = "End Motif Frequency") +
  theme(axis.text.x = element_blank()) +
  #stat_compare_means(aes(label = paste0("P=", ..p.format..)), label.y = 0.155, size = 2)+ #add global p
  stat_compare_means(comparisons = my_comparisons, aes(label = ..p.signif..), label.y = c(0.145, 0.136), ,size = 2.5)


# #Define image directory 
# imagedir="data/cfdna_pipeline/22-summarize_motif"
# 
# #Create file path
# out.plot <- file.path(imagedir, "endmotif.2bp.pre.tiff")
# 
# #Save as tiff
# ggsave(filename = out.plot, plot = bp, device = "tiff", dpi = 300, width = 8, height = 3)
```

# Box Plot for Mono-nucleotide End Motif by SMN Status & Plasma Draw Time 
```{r bp_mono_pre, echo = FALSE}

#Define comparison groups
my_comparisons <- list(c("Pre-SMN (N = 4)", "No Cancer (N = 50)"), 
                       c("Post-SMN (N = 22)", "No Cancer (N = 50)"),
                       c("Pre-SMN (N = 4)", "Post-SMN (N = 22)"))

#Set Color palette
viridis_palette <- viridis(30)
custom_color <- viridis_palette[c(1, 9, 16, 24, 29)]
show_col(viridis_palette)

#Create box plot
bp1.mono.pre <- ggplot(summary.motif.mono.pre, aes(x = group3, y = mono_sum.corrected, fill = group3)) +
  geom_boxplot() +
  scale_fill_manual(name = "Group", values = custom_color)+
  facet_grid(. ~ mono_motif) + 
  labs(x = "Group", y = "End Motif Frequency") +
  theme(axis.text.x = element_blank()) +
  # stat_compare_means(aes(label = paste0("p=", ..p.format..)),
  #                   label.x = 2.4, label.y = 0.4, size = 3) + #global p value
  stat_compare_means(comparisons = my_comparisons, aes(label = ..p.format..), size = 2.5, label.y =c(0.37, 0.355, 0.34))


#Combine both mono- and di-nucleotide plots 
bp.pre.combined <- ggarrange(bp.2bp.pre, bp1.mono.pre, ncol = 1, nrow = 2, labels = c("D", "E"), common.legend = T, legend = "bottom")


```



# Combine all plots
```{r combine_plots, echo = FALSE}



# Adjust layout matrix to have the heatmap span multiple rows.
# For this example, the heatmap will span 4 rows on the left side, and the ggplot2 plots will be on the right.
lm <- matrix(c(1,2), ncol=2) 
labels <- c("A", "")


# Combine box plots from both SMN and SMN/Plasma Draw
p.combined <- ggarrange(p.smn.motif.combined, bp.pre.combined, ncol = 1, nrow = 2, common.legend = FALSE, legend = "none")

# Arrange the plots
combinedplots <- grid.arrange(grobs = list(p1[[4]], p.combined), 
             layout_matrix = lm) 


annotated_plot <- annotate_figure(combinedplots, 
                                  top = text_grob("A", size = 14, face = "bold", x = 0.01, y = -0.5, hjust = 0))

#save file
imagedir="accepted_manuscript"

out.plot <- file.path(imagedir, "Figure3_endmotifsALL.tiff")

#save as tiff
ggsave(filename = out.plot, plot = annotated_plot, device = "tiff", dpi = 300, width = 12, height = 10, bg = "white")


```