# Load library
```{r libraries}
library(tidyverse)
library(ggpubr)
library(rstatix)
library(stringr)
library(RColorBrewer)
library(scales)
library(viridis)
```

# Import End Motif RDS file
```{r import_clean}

#import end nmotif rds file 
df.motif <- readRDS("data/cfdna_pipeline/21-combine_motif/endmotif_2bp_gc_summary.rds")

#summarize counts as dataframe still stratified by gc-strata
df.motif <- df.motif %>%
  group_by(id, end_motif, frac) %>%
  summarise(
    count = sum(count), 
    count.corrected = sum(count.corrected), 
    frac.total = first(frac.total),
    frac.total.corrected = first(frac.total.corrected)
  )
  
#Import meta data
meta <- read_csv("data/sample_reference_sorted230325.csv")
meta <- dplyr::rename(meta, id = Legacy.Custom.ID) #rename to id

#select specific meta data and cancer type
meta <- meta %>% select(id, Cancer, pre_cancer, smn, Breast.Cancer, Thyroid.Cancer) 

#replace NA with "No"
meta <- meta %>%
  mutate_at(vars(smn), ~replace_na(., "No"))

#merge meta data with motif data
summary.motif <- inner_join(meta, df.motif, by="id") 

#create phenotype subgroups
summary.motif <- summary.motif %>% 
  mutate(
    group = case_when(
      Cancer == "Yes" & pre_cancer == "No" ~ "Post-Dx", 
      Cancer == "Yes" & pre_cancer == "Yes" ~ "Pre-Dx",
      Cancer == "No" & pre_cancer == "No" ~ "No Cancer"), 
    group2 = case_when(
      Cancer == "Yes" & smn == "Yes" ~ "SMN",
      Cancer == "Yes" & smn == "No" ~ "PMN",
      Cancer == "No" & smn == "No" ~ "No Cancer"), 
    group3 = case_when(
      id %in% c("CCF02224-01-001", "CCF07535-01-001", "CCF02986-01-001", "CCF06617-01-001") ~ "Pre-SMN",
      id %in% c("CCF02583-01-001", "CCF02068-01-001", "CCF04317-01-001", "CCF07793-01-001", "CCF05875-02-001") ~ "Pre-PMN",
      TRUE ~ group2
    )
  ) %>% 
  mutate(
    ca_type1 = case_when(
      Cancer == "No" ~ "No Cancer",
      Breast.Cancer == "Yes" & Thyroid.Cancer == "Yes" ~ "Both",
      Breast.Cancer == "Yes" & Thyroid.Cancer == "No" ~ "Breast",
      Breast.Cancer == "No" & Thyroid.Cancer == "Yes" ~ "Thyroid"
    )
  )

# Rename levels in the group3 column 
summary.motif <- summary.motif %>%
  mutate(group3 = fct_recode(group3,
                             "Post-PMN" = "PMN",
                             "Post-SMN" = "SMN"))

# Relocate columns 
summary.motif <- summary.motif %>% 
  relocate(Cancer, .after = id) %>% 
  relocate(pre_cancer, .after = Cancer) %>%
  relocate(group, .after = pre_cancer)  %>%
  relocate(ca_type1, .after = group) %>%
  relocate(group2, .after = group) %>%
  relocate(group3, .after = group2)
  ungroup()

#Export as csv
write.csv(summary.motif,"data/cfdna_pipeline/22-summarize_motif/22-endmotif_summary.csv", row.names = FALSE)
```

# Summarise Di-nucleotide End Motif Profiles by Cancer Subgroups 
```{r create_df, echo = FALSE}

summary.motif <- read.csv("data/cfdna_pipeline/22-summarize_motif/22-endmotif_summary.csv")

#End Motif Profile by SMN status
summary.motif.smn <- summary.motif %>% 
  group_by(id, end_motif, group2) %>%
  summarise(
    count = sum(count), 
    count.corrected = sum(count.corrected)
  ) %>% ungroup() %>% 
  group_by(id) %>% 
  mutate(
    total = sum(count),
    total.corrected = sum(count.corrected),
    freq = count / total, 
    freq.corrected = count.corrected / total.corrected
  ) %>%
  mutate(
    group2 = factor(group2, levels = c("SMN", "PMN", "No Cancer"))
  )

# Set the sample sizes for each group
n_smn <- 26
n_pmn <- 23
n_no_cancer <- 50

# Update group names with sample sizes
summary.motif.smn <- summary.motif.smn %>%
  mutate(group2 = recode(group2,
                         `SMN` = paste0("SMN (N = ", n_smn, ")"),
                         `PMN` = paste0("PMN (N = ", n_pmn, ")"),
                         `No Cancer` = paste0("No Cancer (N = ", n_no_cancer, ")")))

#Reshape dataframe using pivot_longer
summary.motif.smn.long <- summary.motif.smn %>%
  pivot_longer(cols = c(freq, freq.corrected),
               names_to = "type",
               values_to = "value")


#End Motif Profile by SMN Status and Draw Time 
summary.motif.pre <- summary.motif %>% 
  group_by(id, end_motif, group3) %>%
  summarise(
    count = sum(count), 
    count.corrected = sum(count.corrected)
  ) %>% ungroup() %>% 
  group_by(id) %>% 
  mutate(
    total = sum(count),
    total.corrected = sum(count.corrected),
    freq = count / total, 
    freq.corrected = count.corrected / total.corrected
  ) %>%
  mutate(group3 = factor(group3, levels = c("Pre-SMN", "Post-SMN", "Pre-PMN", "Post-PMN", "No Cancer")))


#reshapelong 
summary.motif.pre.long <- summary.motif.pre %>%
  pivot_longer(cols = contains("freq"), 
               names_to = "type", 
               values_to = "value") 

# Set the sample sizes for each group
n_smn <- 4
n_postsmn <- 22
n_pmn <- 5
n_postpmn <- 18
n_no_cancer <- 50

# Update group names with sample sizes
summary.motif.pre <- summary.motif.pre %>%
  mutate(group3 = recode(group3,
                         `Pre-SMN` = paste0("Pre-SMN (N = ", n_smn, ")"),
                         `Pre-PMN` = paste0("Pre-PMN (N = ", n_pmn, ")"),
                         `Post-SMN` = paste0("Post-SMN (N = ", n_postsmn, ")"),
                         `Post-PMN` = paste0("Post-PMN (N = ", n_postpmn, ")"),
                         `No Cancer` = paste0("No Cancer (N = ", n_no_cancer, ")")))
```
# Summarise Mono-nucleotide End Motif Profiles by Cancer Subgroups 
```{r create_df_mono, echo = FALSE}

#Mono motif by SMN status 
summary.motif.smn.mono <- summary.motif.smn %>%
  #obtain mono-nucleotide motifs 
  mutate(mono_motif = str_extract(end_motif, "\\b[A-Z]{1}")) %>%
  group_by(id, mono_motif, group2) %>%
  summarise(
    freq = sum(freq),
    freq.corrected = sum(freq.corrected)
  ) %>%
  mutate(
    group2 = factor(group2, levels = c("SMN", "PMN", "No Cancer"))
  )

#Reshape using pivot_longer 
summary.motif.frac.long <- summary.motif.smn.mono %>%
  pivot_longer(cols = contains("freq"), 
               names_to = "type", 
               values_to = "value") 

#Mono motifs by SMN Status and Draw Time 
summary.motif.mono.pre <- summary.motif.pre %>%
 mutate(mono_motif = str_extract(end_motif, "\\b[A-Z]{1}")) %>%
  group_by(id, mono_motif, group3) %>%
  summarise(
    mono_sum = sum(freq),
    mono_sum.corrected = sum(freq.corrected)
  )

#Reshape using pivot_longer 
summary.motif.mono.pre.long <- summary.motif.mono.pre %>%
  pivot_longer(cols = contains("mono_sum"), 
               names_to = "type", 
               values_to = "value") 
```
# Box plot for Di-nucleotide End Motif by SMN Status 
```{r bp_dismn, echo = FALSE}

#Define comparison groups 
my_comparisons <- list(c("SMN (N = 26)", "No Cancer (N = 50)"), c("PMN (N = 23)", "No Cancer (N = 50)"))

#Set color palette 
viridis_palette <- viridis(30)
custom_color <- viridis_palette[c(1, 16, 29)]
show_col(viridis_palette)


#Create box plot facetted by end-motif
bp.smn <- ggplot(summary.motif.smn, aes(x = group2, y = freq.corrected, fill = group2)) + 
  geom_boxplot() +
  facet_grid(. ~ end_motif) + 
  scale_fill_manual(name = "Group", values = custom_color) + # Use custom colors
  labs(x = "Group", y = "End Motif Frequency") +
  theme(axis.text.x = element_blank()) +
  #stat_compare_means(aes(label = paste0("P=", ..p.format..)), label.y = 0.155, size = 2) + #add global p
  stat_compare_means(comparisons = my_comparisons, aes(label = ..p.signif..), label.y = c(0.15, 0.143), size = 2.5)

#Define image directory 
imagedir="data/cfdna_pipeline/22-summarize_motif"

#Create file path
out.plot <- file.path(imagedir, "endmotif.2bp.smn.tiff")

#Save as tiff
ggsave(filename = out.plot, plot = bp, device = "tiff", dpi = 300, width = 8, height = 3)

```

# Box Plot for Di-nucleotide by SMN Status & Plasma Draw Time
```{r bp_dipre, echo = FALSE}

#Define comparison groups
my_comparisons <- list(c("Pre-SMN (N = 4)", "No Cancer (N = 50)"), 
                       c("Pre-SMN (N = 4)", "Post-SMN (N = 22)"))

#Set color palette 
viridis_palette <- viridis(30)
custom_color <- viridis_palette[c(1, 9, 16, 24, 29)]
show_col(viridis_palette)

#Create box plot facetted by end-motif
bp.2bp.pre <- ggplot(summary.motif.pre, aes(x = group3, y = freq.corrected, fill = group3)) + 
  geom_boxplot() +
  facet_grid(. ~ end_motif) + 
  scale_fill_manual(name = "Group", values = custom_color)+
  labs(x = "Group", y = "End Motif Frequency") +
  theme(axis.text.x = element_blank()) +
  #stat_compare_means(aes(label = paste0("P=", ..p.format..)), label.y = 0.155, size = 2)+ #add global p
  stat_compare_means(comparisons = my_comparisons, aes(label = ..p.signif..), label.y = c(0.145, 0.136), ,size = 2.5)


#Define image directory 
imagedir="data/cfdna_pipeline/22-summarize_motif"

#Create file path
out.plot <- file.path(imagedir, "endmotif.2bp.pre.tiff")

#Save as tiff
ggsave(filename = out.plot, plot = bp, device = "tiff", dpi = 300, width = 8, height = 3)
```

# Box Plot for Mono-nucleotide End Motif by SMN Status
```{r bp_monosmn, echo = FALSE}

#Define comparison groups
my_comparisons <- list(c("SMN", "No Cancer"), c("SMN", "PMN"), c("PMN", "No Cancer"))

#Set color palette 
viridis_palette <- viridis(30)
custom_color <- viridis_palette[c(1, 16, 29)]
show_col(viridis_palette)


#Create box plot 
bp.mono.smn <- ggplot(summary.motif.smn.mono, aes(x = group2, y = freq.corrected, fill = group2)) + 
  geom_boxplot() +
  scale_fill_manual(name = "Group", values = custom_color) + # Use custom colors
  facet_grid(. ~ mono_motif) + 
  labs(x = "Group", y = "End Motif Frequency") +
  theme(axis.text.x = element_blank()) +
  # stat_compare_means(aes(label = paste0("p=", ..p.format..)),
  #                    label.x = 1.75, label.y = 0.4, size = 3) + #global p value
  stat_compare_means(comparisons = my_comparisons, aes(label = ..p.format..), size = 2.5, label.y =c(0.37, 0.355, 0.345))

#Combine mono- and di-nucleotide end motif plots together 
p.smn.motif.combined <- ggarrange(bp.mono.smn, bp.smn, ncol = 1, nrow = 2, labels = c("A", "B"), common.legend = TRUE, legend = "bottom")


#write for-loop to save files 
imagedir="data/cfdna_pipeline/22-summarize_motif"

#Define directory 
out.plot <- file.path(imagedir, "endmotif_smn_monodi_combined.tiff")

#Save as tiff
ggsave(filename = out.plot, plot = p.smn.motif.combined, device = "tiff", dpi = 300, width = 8, height = 6)

```

# Box Plot for Mono-nucleotide End Motif by SMN Status & Plasma Draw Time 
```{r bp_monosmn, echo = FALSE}

#Define comparison groups
my_comparisons <- list(c("Pre-SMN (N = 4)", "No Cancer (N = 50)"), 
                       c("Post-SMN (N = 22)", "No Cancer (N = 50)"),
                       c("Pre-SMN (N = 4)", "Post-SMN (N = 22)"))

#Set Color palette
viridis_palette <- viridis(30)
custom_color <- viridis_palette[c(1, 9, 16, 24, 29)]
show_col(viridis_palette)

#Create box plot
bp1.mono.pre <- ggplot(summary.motif.mono.pre, aes(x = group3, y = mono_sum.corrected, fill = group3)) +
  geom_boxplot() +
  scale_fill_manual(name = "Group", values = custom_color)+
  facet_grid(. ~ mono_motif) + 
  labs(x = "Group", y = "End Motif Frequency") +
  theme(axis.text.x = element_blank()) +
  # stat_compare_means(aes(label = paste0("p=", ..p.format..)),
  #                   label.x = 2.4, label.y = 0.4, size = 3) + #global p value
  stat_compare_means(comparisons = my_comparisons, aes(label = ..p.format..), size = 2.5, label.y =c(0.37, 0.355, 0.34))


#Combine both mono- and di-nucleotide plots 
bp.pre.combined <- ggarrange(bp1.mono.pre, bp.2bp.pre, ncol = 1, nrow = 2, labels = c("A", "B"), common.legend = T, legend = "bottom")

#write for-loop to save files 
imagedir="data/cfdna_pipeline/22-summarize_motif"

#Define directory
out.plot <- file.path(imagedir, "endmotif_pre_monodi_combined.tiff")

#save as tiff
ggsave(filename = out.plot, plot = bp.pre.combined, device = "tiff", dpi = 300, width = 8, height = 6)
```

