---
title: "cfDNA fragment end motif analysis and visualization"
author: "Darren Liu"
date: "2024-03-04"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = "W:/Shared/LRI/Labs/engclab/LAB_MEMBER_CENTRAL/Darren_Liu/3_cf-DNA")

```

# Load library
```{r libraries}

library(tidyverse)
library(ggpubr)
library(rstatix)
library(RColorBrewer)
library(scales)
library(viridis)
library(pheatmap)

```
# Import Data
```{r import}

# Import end motif and meta data  
df.motif <- readRDS("protocol_ms/data/21-combine_motif/endmotif_4bp_gc_summary.rds")
meta <- read.csv("protocol_ms/data/phtscfx_meta240304.csv")

# Merge data 
df.motif <- inner_join(meta, df.motif, by="id") 
```

# Calculate and summarize end motif frequencies
```{r summarize}

# Calculate total count for each end motif  
df.motif <- df.motif %>% 
  group_by(id, end_motif, group2, group3, ca_1st, ca_2nd) %>%
  summarise(
    count = sum(count), 
    count.corrected = sum(count.corrected)
  ) %>% 
  ungroup()

# Calculate end motif frequency 
df.motif <- df.motif %>% 
  group_by(id) %>% 
  mutate(
    total = sum(count),
    total.corrected = sum(count.corrected),
    freq = count / total, 
    freq.corrected = count.corrected / total.corrected
  ) %>%
  mutate(
    group2 = factor(group2, levels = c("SMN", "PMN", "No Cancer")),
    group3 = factor(group3, levels = c("Pre-SMN", "Post-SMN", "Pre-PMN", "Post-PMN", "No Cancer"))
  ) %>% 
  ungroup() 


# Save as csv
#write.csv(df.motif, "protocol_ms/data/22-motif_analysis/22-motif_analysis.4bp.csv", row.names = FALSE)
#saveRDS(df.motif, "protocol_ms/data/22-motif_analysis/22-motif_analysis.4bp.rds")
```

# Statistical analysis 
```{r stat}

## Perform overall testing using KW 
# By SMN status
res.smn.kw <- df.motif %>%
  group_by(end_motif) %>%
  kruskal_test(freq.corrected ~ group2)

# By SMN status and plasma draw time 
res.pre.kw <- df.motif %>%
  group_by(end_motif) %>%
  kruskal_test(freq.corrected ~ group3)

# Filter for only statistically significant
sig.smn.4mer <- res.smn.kw %>% filter(p < 0.05)
sig.pre.4mer <- res.pre.kw %>% filter(p < 0.05)

## Post-hoc pairwise comparison 
# By SMN status
res.smn.wx <- df.motif %>%
  filter(end_motif %in% sig.smn.4mer$end_motif) %>%
  group_by(end_motif) %>%
  wilcox_test(freq.corrected ~ group2) %>%
  adjust_pvalue(method = "BH") %>%
  add_significance() %>%
  arrange(p)

# By SMN status and plasma draw time 
res.pre.wx <- df.motif %>%
  filter(end_motif %in% sig.pre.4mer$end_motif) %>%
  group_by(end_motif) %>%
  wilcox_test(freq.corrected ~ group3) %>%
  adjust_pvalue(method = "BH") %>%
  add_significance() %>%
  arrange(p)

# Save files 
write.csv(res.smn.kw, "protocol_ms/data/22-motif_analysis/res_4bp_smn_kw.csv")
write.csv(res.pre.kw, "protocol_ms/data/22-motif_analysis/res_4bp_pre_kw.csv")
write.csv(res.smn.wx, "protocol_ms/data/22-motif_analysis/res_4bp_smn_wx.csv")
write.csv(res.pre.wx, "protocol_ms/data/22-motif_analysis/res_4bp_pre_wx.csv")
```

Figure 4A. Heatmap 4-mer end motifs 
```{r pheatmap}

# Combine the dataframes
sig.4mer <- rbind(sig.pre.4mer, sig.smn.4mer)

# Remove duplicates based on the end_motif column
sig.4mer <- sig.4mer %>% select(end_motif) %>% distinct()

# Subset significant 4-mers
df.motif.sig <- df.motif %>% filter(end_motif %in% sig.4mer$end_motif)

# Pivot the data to wide format
df.motif.sig <- df.motif.sig %>%
  select(id, end_motif, freq.corrected) %>%
  pivot_wider(names_from = id, values_from = freq.corrected)

# Convert to matrix 
heatmap_matrix <- as.matrix(df.motif.sig[,-1])
rownames(heatmap_matrix) <- df.motif.sig$end_motif

# Create annotations
annotation_data <- df.motif %>%
  select(id, group2, group3, ca_1st, ca_2nd) %>% 
  mutate_at(vars(ca_1st, ca_2nd), ~replace_na(., "None")) %>% 
  distinct() %>%
  arrange(id) %>%
  column_to_rownames("id") 


# Color palette using viridis
viridis_palette <- viridis(30) # Define palette
show_col(viridis_palette) # Examine colors 
colors <- viridis_palette[c(1, 9, 16, 24, 29)] # Select colors

# Define color mapping for patient subgroups 
group2_colors <- c("SMN" = colors[1], "PMN" = colors[3], "No Cancer" = colors[5])
group3_colors <- c("Pre-SMN" = colors[1], "Post-SMN" = colors[2], 
                   "Pre-PMN" = colors[3], "Post-PMN" = colors[4], 
                   "No Cancer" = colors[5])

# Define color mapping for cancer type 
all_cancers <- unique(c(annotation_data$ca_1st, annotation_data$ca_2nd))
cancer_colors <- brewer.pal(length(all_cancers), "Dark2")
color_mapping <- setNames(cancer_colors, all_cancers)

# Rename annotations labels
names(annotation_data)[names(annotation_data) == "group2"] <- "SMN"
names(annotation_data)[names(annotation_data) == "group3"] <- "Draw Time"
names(annotation_data)[names(annotation_data) == "ca_1st"] <- "1st Ca."
names(annotation_data)[names(annotation_data) == "ca_2nd"] <- "2nd Ca."

# Define list of annotation category with color mapping
annotation_colors <- list(
    SMN = group2_colors,         # Changed 'group2' to 'SMN'
    `Draw Time` = group3_colors, # Changed 'group3' to 'Draw Time'
    `1st Ca.` = color_mapping,   # Changed 'ca_1st' to '1st Ca.'
    `2nd Ca.` = color_mapping    # Changed 'ca_2nd' to '2nd Ca.'
)

# Generate heat map 
p1 <- pheatmap(heatmap_matrix,
         scale = "row",
         cluster_rows = TRUE, 
         cluster_cols = TRUE, 
         show_rownames = TRUE, 
         show_colnames = FALSE, 
         border_color = NA, 
         treeheight_row = 0, 
         #treeheight_col = 0,
         color = viridis::viridis(120, option = "magma"),
         #height = 8,
         #width = 8, 
         annotation_col = annotation_data,
         annotation_colors = annotation_colors)

# Save heatmap 
ggsave(filename = "protocol_ms/data/22-motif_analysis/end_motif4bp.tiff", plot = p1, device = "tiff", dpi = 300, width = 4, height = 8)

```

Figure 4B. Boxplot 1-mer by SMN status 
```{r 1mer_smn}

# Function to extract the first n characters from 4-mer end motifs
calculate_nmer <- function(df, group_var) {
  df %>% 
    mutate(end_motif = str_extract(end_motif, sprintf("\\b[A-Z]{%d}", n))) %>% 
    group_by(id, end_motif, .data[[group_var]]) %>%
    summarise(
      freq = sum(freq), 
      freq.corrected = sum(freq.corrected), 
      .groups = "drop"
    )
}

# Calcualte n-mer 
n <- 1
df.motif_nmer <- calculate_nmer(df.motif, "group2")

# Select color palette for patient subgroups 
viridis_palette <- viridis(30)
show_col(viridis_palette)
custom_color <- viridis_palette[c(1, 16, 29)]  

# Define pairwise comparison groups
my_comparisons <- list(c("SMN", "No Cancer"), c("PMN", "No Cancer")) 


# Generate boxplot
bp.nmer <- ggplot(df.motif_nmer, aes(x = group2, y = freq.corrected, fill = group2)) +
  geom_boxplot() +
  scale_fill_manual(name = "Group", values = custom_color) + # Use custom colors
  facet_grid(. ~ end_motif) + 
  labs(x = "Group", y = "End Motif Frequency") +
  theme(axis.text.x = element_blank()) +
  stat_compare_means(comparisons = my_comparisons, aes(label = ..p.format..), size = 2.5, label.y =c(0.37, 0.355, 0.345)) # Adjust placement of p-value labels


# Save boxplot  
ggsave(filename = "protocol_ms/data/22-motif_analysis/end_motif_1bp_smn.tiff", plot = bp.nmer, device = "tiff", dpi = 300, width = 8, height = 3)

```

Figure 4C. Boxplot 2-mer by SMN status 
```{r 2mer_smn}

# Calcualte n-mer 
n <- 2
df.motif_nmer <- calculate_nmer(df.motif, "group2")

# Select color palette for patient subgroups 
viridis_palette <- viridis(30)
show_col(viridis_palette)
custom_color <- viridis_palette[c(1, 16, 29)] 

# Define pairwise comparison groups
my_comparisons <- list(c("SMN", "No Cancer"), c("PMN", "No Cancer")) 

# Generate boxplot
bp.nmer <- ggplot(df.motif_nmer, aes(x = group2, y = freq.corrected, fill = group2)) +
  geom_boxplot() +
  scale_fill_manual(name = "Group", values = custom_color) + # Use custom colors
  facet_grid(. ~ end_motif) + 
  labs(x = "Group", y = "End Motif Frequency") +
  theme(axis.text.x = element_blank()) +
  stat_compare_means(comparisons = my_comparisons, aes(label = ..p.signif..), label.y = c(0.15, 0.143), size = 2.5) # Adjust significance labels  


# Save boxplot  
ggsave(filename = "protocol_ms/data/22-motif_analysis/end_motif_2bp_smn.tiff", plot = bp.nmer, device = "tiff", dpi = 300, width = 8, height = 3)

```

# Figure 4D. Boxplot 1-mer by SMN status & plasma draw timing
```{r 1mer_pre}

# Calcualte n-mer 
n <- 1
df.motif_nmer <- calculate_nmer(df.motif, "group3")

# Select color palette for patient subgroups 
viridis_palette <- viridis(30)
show_col(viridis_palette)
custom_color <- viridis_palette[c(1, 9, 16, 24, 29)] 

# Define pairwise comparison groups
my_comparisons <- list(c("Pre-SMN", "No Cancer"), 
                       c("Post-SMN", "No Cancer"),
                       c("Pre-SMN", "Post-SMN"))


# Generate boxplot
bp.nmer <- ggplot(df.motif_nmer, aes(x = group3, y = freq.corrected, fill = group3)) +
  geom_boxplot() +
  scale_fill_manual(name = "Group", values = custom_color) + # Use custom colors
  facet_grid(. ~ end_motif) + 
  labs(x = "Group", y = "End Motif Frequency") +
  theme(axis.text.x = element_blank()) +
  stat_compare_means(comparisons = my_comparisons, aes(label = ..p.format..), size = 2.5, label.y =c(0.37, 0.355, 0.34)) # Adjust placement of p-value labels


# Save boxplot  
ggsave(filename = "protocol_ms/data/22-motif_analysis/end_motif_1bp_pre.tiff", plot = bp.nmer, device = "tiff", dpi = 300, width = 8, height = 3)

```

# Figure 4E. Boxplot 2-mer by SMN status & plasma draw timing 
```{r 2mer_pre}

# Calcualte n-mer 
n <- 2
df.motif_nmer <- calculate_nmer(df.motif, "group3")

# Select color palette for patient subgroups 
viridis_palette <- viridis(30)
show_col(viridis_palette)
custom_color <- viridis_palette[c(1, 9, 16, 24, 29)] 

# Define pairwise comparison groups
my_comparisons <- list(c("Pre-SMN", "No Cancer"), c("Pre-SMN", "Post-SMN")) 


# Generate boxplot
bp.nmer <- ggplot(df.motif_nmer, aes(x = group3, y = freq.corrected, fill = group3)) +
  geom_boxplot() +
  scale_fill_manual(name = "Group", values = custom_color) + # Use custom colors
  facet_grid(. ~ end_motif) + 
  labs(x = "Group", y = "End Motif Frequency") +
  theme(axis.text.x = element_blank()) +
  stat_compare_means(comparisons = my_comparisons, aes(label = ..p.signif..), label.y = c(0.145, 0.136), size = 2.5) # Adjust significance labels  

# Save boxplot  
ggsave(filename = "protocol_ms/data/22-motif_analysis/end_motif_2bp_pre.tiff", plot = bp.nmer, device = "tiff", dpi = 300, width = 8, height = 3)
```
Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
